---
title: "Why the catch size distribution does not pin down the parameters"
author: "Gustav Delius"
format:
  html:
    toc: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Why the catch size distribution does not pin down the parameters}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.2)
```

## The question

`matchCatch()` adjusts gear selectivity, catchability, mortality and the
growth-diffusion coefficient so that the model reproduces an observed catch
size distribution. A natural hope is that fitting the catch recovers the
*true* values of those parameters. It does not, and it cannot — not from
the catch shape alone.

This is not a numerical accident or a weakness of the optimiser. It is a
structural property of the problem. The test
`"matchCatch recovers a model-generated catch size distribution"` checks
only that the *fitted catch distribution* matches the observed one, and its
comment warns that "the gear and mortality parameters themselves are not
uniquely recovered". This vignette builds intuition for *why*, through
examples, and looks in particular at the case that matters most in
practice: **the total yield is known but the production is not.**

## The catch is a product, but the population is not free

The model's predicted catch density at weight $w$ is

$$
C(w) \;=\; N(w)\, F(w),
\qquad F(w) = q\, S(w),
$$

where $N(w)$ is the steady-state number density, $S(w)$ the gear
selectivity (between 0 and 1) and $q$ the catchability. This is exactly
what the package computes — in `plot_catch()` the modelled catch is
`initial_n * getFMort`.

It would be glib to say "the catch is a product, so you cannot recover its
factors", because the population $N$ is **not** a free factor we may choose
at will. In the steady state, $N(w)$ is determined by the biology: it is
the solution of the size-structured balance between

- **growth** $g(w)$ — the deterministic rate at which an individual gains
  weight,
- **diffusion** $d(w) = D_{\text{ext}}\, w^{\,n+1}$ — the stochastic spread
  in growth, so that individuals of the same age reach a *range* of sizes,
- **mortality** $\mu(w)$ — which removes individuals and steepens the
  spectrum.

The real question is therefore sharper: given the catch shape, can we
disentangle *selectivity* from the *biological parameters that shape the
spectrum* — chiefly mortality and diffusion? The answer is no, and the
reason is that **mortality and diffusion both reshape the upper tail of
$N(w)$, so the catch shape sees only their combined effect.**

## A look at the model

Throughout we use a single-gear Celtic Sea model and generate catch data
from it, so that the truth is known and we can ask whether fitting the
catch recovers it. The helpers below build that model, sample its predicted
catch into 1-cm length bins, and let us set the diffusion coefficient.

```{r}
#| message: false
#| warning: false
library(mizerEcopath)

# Collapse celtic_params to one gear per species.
make_single_gear <- function() {
    params <- celtic_params
    gp <- gear_params(params)
    keep <- do.call(rbind, lapply(split(gp, gp$species), function(g) {
        row <- g[which.max(g$yield_observed), , drop = FALSE]
        row$gear <- "total"
        row$yield_observed <- sum(g$yield_observed)
        row
    }))
    rownames(keep) <- paste(keep$species, keep$gear, sep = ", ")
    gear_params(params) <- keep
    initial_effort(params) <- 1
    params
}

# Sample the model's own predicted catch into 1-cm length bins.
model_catch_data <- function(params, species, n_total = 1e5) {
    sp <- species_params(params); s <- sp$species == species; sps <- sp[s, ]
    lengths <- (params@w / sps$a)^(1 / sps$b)
    l_max <- (sps$w_max / sps$a)^(1 / sps$b)
    rate <- params@initial_n[s, ] * getFMort(params)[s, ] * params@dw
    df <- aggregate(rate, list(length = floor(lengths)), sum)
    names(df)[2] <- "rate"
    df <- df[df$rate > 0 & (df$length + 1) <= l_max, ]
    count <- round(df$rate / sum(df$rate) * n_total)
    data.frame(species = species, gear = "total", length = df$length[count > 0],
               dl = 1, catch = count[count > 0])
}

# The model's predicted catch density over the observed length bins.
model_catch_pdf <- function(params, species, bins) {
    sp <- species_params(params); s <- sp$species == species; sps <- sp[s, ]
    lengths <- (params@w / sps$a)^(1 / sps$b)
    rate <- params@initial_n[s, ] * getFMort(params)[s, ] * params@dw
    h <- tapply(rate, factor(floor(lengths), levels = bins), sum)
    h[is.na(h)] <- 0; as.numeric(h / sum(h))
}

# Set the diffusion coefficient D_ext and recompute the (diffusion-aware)
# steady state, holding mortality, growth and selectivity fixed.
set_D_ext <- function(params, species, D) {
    sp <- species_params(params); s <- sp$species == species
    sp$D_ext[s] <- D; species_params(params) <- sp
    params@ext_diffusion[s, ] <- D * params@w^(sp$n[s] + 1)
    steadySingleSpecies(params, species = species)
}

p0      <- make_single_gear()
species <- "Haddock"
s       <- species_params(p0)$species == species
sps     <- species_params(p0)[s, ]
lengths <- (p0@w / sps$a)^(1 / sps$b)   # length (cm) of each weight bin
```

## How diffusion shapes the spectrum

Diffusion is the parameter readers tend to have least intuition for, so it
is worth pinning down with a measurement.

Without diffusion ($D_{\text{ext}} = 0$) every individual of a cohort
follows the same deterministic growth curve. That curve slows to a halt at
an *asymptotic size*, well short of the maximum size $w_{\max}$, because a
mature fish channels a growing share of its surplus energy into
reproduction rather than growth. For this Haddock model the asymptotic
length is about 56 cm, whereas $w_{\max}$ corresponds to 112 cm. With no
diffusion, then, *no fish ever exceeds 56 cm*: the spectrum has a hard edge
there and the upper half of the size range is simply empty.

Diffusion lets growth rates vary between individuals: a cohort *spreads
out* into a range of sizes as it ages (this is what the age-fitting
machinery in `project_diffusion()` exploits; see the *Diffusion*
vignette). Crucially, it lets some individuals grow *past* the
deterministic asymptote and on towards $w_{\max}$. The larger the
diffusion, the more individuals populate that upper tail. So **more
diffusion means *more* large fish** — it is diffusion, not deterministic
growth, that fills the large-size end of the spectrum.

We can see it directly in the steady-state spectra. Holding mortality and
selectivity fixed, we recompute the spectrum at three diffusion levels and
rescale each to the **same total biomass**, so the comparison is fair: any
difference is then in how that fixed biomass is *distributed* over sizes,
not in how much there is. We plot biomass density, so the area under each
curve is the (equal) total biomass. At low diffusion the biomass is bottled
up below the deterministic asymptote (a sharp edge near 56 cm); raising the
diffusion spreads it out into the large-size tail:

```{r}
#| message: false
#| warning: false
#| fig-cap: "Steady-state biomass density at three diffusion coefficients, each rescaled to the same total biomass (equal area under the curve), with mortality and selectivity held fixed. At low `D_ext` no fish pass the ~56 cm deterministic asymptote; larger `D_ext` fills the large-size tail. More diffusion means more large fish."
l_mat <- (sps$w_mat / sps$a)^(1 / sps$b)   # length at maturity (cm)
l_max <- (sps$w_max / sps$a)^(1 / sps$b)   # maximum length (cm)
dl   <- p0@dw / (sps$a * sps$b * lengths^(sps$b - 1))  # width of each bin (cm)
Ds   <- c(0.1, 10, 100)
cols <- c("blue", "black", "red")
rng  <- lengths >= 0.5 * l_mat & lengths <= 0.97 * l_max

# Biomass density in length, rescaled so the total biomass equals 1.
biomass_density <- function(D) {
    pp <- set_D_ext(p0, species, D)
    Bm <- pp@initial_n[s, ] * p0@w * p0@dw   # biomass in each bin
    (Bm / dl) / sum(Bm)
}
beta <- lapply(Ds, biomass_density)
ymax <- max(vapply(beta, function(b) max(b[rng]), numeric(1)))

plot(NULL, xlim = c(0.5 * l_mat, 0.97 * l_max), ylim = c(0, ymax),
     xlab = "Length [cm]", ylab = "Biomass density (total biomass = 1)",
     main = "Biomass spectrum at equal total biomass")
for (i in seq_along(Ds)) lines(lengths[rng], beta[[i]][rng], col = cols[i], lwd = 2.5)
abline(v = l_mat, lty = 3); text(l_mat, ymax, " maturity", adj = c(0, 1), cex = 0.8)
legend("topright", paste0("D_ext = ", Ds), col = cols, lwd = 2.5, bty = "n")
```

The effect on the catch follows directly. Raising `D_ext` lets the catch
reach larger sizes: with little diffusion almost nothing is caught above
60 cm, while at high diffusion a sixth of the catch is, and the mean catch
length climbs:

```{r}
#| message: false
#| warning: false
diff_demo <- do.call(rbind, lapply(c(1e-3, 0.1, 1, 10, 100), function(D) {
    pp <- set_D_ext(p0, species, D)
    cw <- pp@initial_n[s, ] * getFMort(pp)[s, ] * pp@dw
    cw <- cw / sum(cw)                       # normalised catch over weight
    data.frame(D_ext                = D,
               frac_catch_over_60cm = round(sum(cw[lengths > 60]), 3),
               mean_catch_length_cm = round(sum(cw * lengths), 1))
}))
print(diff_demo, row.names = FALSE)
```

Mortality and diffusion therefore push the large-fish abundance in
*opposite* directions: more mortality removes large fish, more diffusion
supplies them. That is the root of the degeneracy. To keep the catch
unchanged while *raising* mortality (which removes large fish), the fit can
simply *raise* the diffusion (which puts them back), and vice versa.
Because the catch reveals only the net upper-tail shape (filtered through
selectivity), it cannot say how much of that shape is mortality and how
much is diffusion. The optimiser slides freely along the trade-off, with
the two rising and falling together.

## Yield known, production unknown

This is the situation we usually face: the total landed weight of a stock
is recorded, but an independent estimate of its production is not. What can
we then learn, and what stays out of reach?

We hold mortality (`mu_mat`) fixed at a range of values spanning a factor
of four around the truth, and at each one fit the catch shape **and** the
yield, but supply **no** production penalty (`production_lambda = 0`). The
gear selectivity, catchability and diffusion are free. If mortality were
identifiable from the catch and yield, only the true value would fit well.

```{r}
#| message: false
#| warning: false
catch_obs <- model_catch_data(p0, species)
obs_pdf   <- catch_obs$catch / sum(catch_obs$catch)
mu0       <- sps$mu_mat

mus <- c(0.5, 0.75, 1, 1.5, 2) * mu0
fits <- lapply(mus, function(mu) {
    pp <- p0
    sp <- species_params(pp); sp$mu_mat[s] <- mu; species_params(pp) <- sp
    suppressWarnings(matchCatch(
        pp, species = species, catch = catch_obs,
        yield_lambda = 1, production_lambda = 0,           # yield only
        map = list(m = factor(NA), mu_mat = factor(NA))))  # mortality fixed
})

summary_tab <- do.call(rbind, Map(function(mu, fit) {
    gpf <- gear_params(fit); gpf <- gpf[gpf$species == species, ]
    spf <- species_params(fit)[s, ]
    pm  <- model_catch_pdf(fit, species, catch_obs$length)
    tv  <- 0.5 * sum(abs(obs_pdf - pm))
    nll <- -sum(catch_obs$catch * log(pmax(pm, 1e-300)))  # catch-shape NLL
    data.frame(
        mu_mat       = round(mu, 3),
        l50          = round(gpf$l50, 1),
        catchability = round(gpf$catchability, 3),
        D_ext        = signif(spf$D_ext, 3),
        TV_distance  = round(tv, 4),
        NLL          = nll,
        yield        = signif(getYield(fit)[s], 3),
        production   = signif(getProduction(fit)[species], 3))
}, mus, fits))
# Show the catch-shape NLL relative to the best-fitting row; the sample size
# is the same in every row, so these differences are directly comparable.
summary_tab$NLL <- round(summary_tab$NLL - min(summary_tab$NLL), 1)
print(summary_tab, row.names = FALSE)
```

Three things stand out, and together they tell the whole story.

**The yield fixes the catchability.** Across the whole range the matched
yield sits on the observed value and the catchability barely moves. The
total landed mass $\int C(w)\,w\,dw = q\!\int S(w)\,N(w)\,w\,dw$ depends
directly on $q$, so requiring the model to reproduce the observed yield
determines $q$ — the one thing the *shape* alone leaves undetermined. This
is the part of the problem that yield data solve cleanly.

**The catch shape barely constrains the mortality.** Every row reproduces
the catch distribution closely — the total-variation distances are all tiny
— even though the assumed mortality ranges over a factor of four. The
fitted catch curves lie almost on top of one another:

```{r}
#| fig-cap: "Catch densities fitted at mortality assumptions spanning a factor of four, with the yield matched but no production constraint. The observed catch (grey bars) is reproduced almost equally well by all of them."
heights <- obs_pdf
plot(NULL, xlim = range(catch_obs$length), ylim = c(0, max(heights) * 1.05),
     xlab = "Length [cm]", ylab = "Relative catch density",
     main = paste(species, "— one catch, a range of mortalities"))
rect(catch_obs$length - 0.5, 0, catch_obs$length + 0.5, heights,
     col = "grey85", border = NA)
cols <- c("blue", "cyan4", "black", "orange", "red")
for (i in seq_along(fits)) {
    lines(catch_obs$length, model_catch_pdf(fits[[i]], species, catch_obs$length),
          col = cols[i], lwd = 2)
}
legend("topright", legend = paste0("mu_mat = ", round(mus, 2)),
       col = cols, lwd = 2, bty = "n")
```

The `TV_distance` column is the total-variation distance between the
observed and fitted catch — a readable measure of shape mismatch. The `NLL`
column is the catch-shape negative log-likelihood, the quantity
`matchCatch()` actually minimises, shown relative to the best-fitting row.
(Its absolute level, around 321,000, is a large constant that every row
carries; only the differences matter, and since the sample size is the same
in every row they are directly comparable.) The likelihood tells a
one-sided story that the uniformly tiny total-variation distances hide. It
is essentially flat at and below the true mortality — over the lower half
of the range it varies by less than three units, and is in fact minimised a
little *below* the truth — but it climbs by tens of units *above* it. So
the catch can rule out grossly excessive mortality, yet it cannot
distinguish the true value from any lower one. The entire lower half of the
range is left for the yield and production to resolve.

**Diffusion is what absorbs the difference.** Look at the `D_ext` column:
as the assumed mortality *rises*, the fitted diffusion *rises* with it.
This is exactly the trade-off of the previous section. A higher mortality
removes large fish; the fit simply raises the diffusion to supply more, so
the upper tail of the catch is restored. The two move *together*, sliding
along a valley in which catch and yield stay almost constant — at least on
the low-mortality side, where, as we just saw, even the likelihood cannot
tell the fits apart.

Note how *mild* the diffusion adjustment is here — `D_ext` moves only from
about 11 to 14 across the whole four-fold range of mortality, not over
orders of magnitude. The
trade-off is gentle precisely because, for this stock, the externally
imposed mortality that we are tuning is a small part of the story: at
maturity it contributes only about 0.06 of a total mortality near 0.29, the
rest being predation. Changing it therefore barely perturbs the spectrum,
which is another way of saying the catch can tell us very little about it.

## The one quantity that does respond: production

The last column is the one that moves with mortality. The **production** —
the rate at which biomass is removed by mortality — climbs steadily across
the fits, while the yield stays pinned to its observed value:

```{r}
#| fig-cap: "With the yield matched but production unconstrained, the fitted yield (open circles) is flat across mortality assumptions, while the implied production (filled circles) rises. Yield carries no information about mortality; production carries some."
op <- par(mar = c(4, 4, 2, 4))
plot(summary_tab$mu_mat, summary_tab$production, type = "b", pch = 19,
     xlab = "assumed mu_mat", ylab = "production", main = "")
par(new = TRUE)
plot(summary_tab$mu_mat, summary_tab$yield, type = "b", pch = 1, lty = 2,
     axes = FALSE, xlab = "", ylab = "", ylim = c(0, max(summary_tab$yield) * 1.3))
axis(4); mtext("yield", side = 4, line = 2.5)
legend("topleft", c("production", "yield"), pch = c(19, 1), lty = c(1, 2),
       bty = "n")
par(op)
```

So if you calibrate to **yield only**, the result is:

| Quantity | Status after a yield-only fit |
|----------|-------------------------------|
| Gear selectivity ($l_{50}, l_{25}$) | Well determined by the catch shape |
| Catchability $q$ | Well determined by the yield |
| Mortality (`mu_mat`) | **Weakly constrained** — bounded from above by the catch, but free below; slides against diffusion |
| Diffusion (`D_ext`) | **Weakly constrained** — slides against mortality |
| Production | Follows the loosely-constrained mortality |

The model you get back will reproduce the observed catch and yield
faithfully, and its selectivity and catchability can be trusted. But its
mortality and diffusion are one point on a shallow valley along which the
catch and yield hardly change, not a measurement. Production moves along
that valley, so it is the natural quantity to bring in if you want to pin
the mortality down.

## What closes the gap

Adding an observed production to the objective (`production_lambda > 0`)
gives the optimiser the one constraint that varies along the valley, so it
selects the mortality — and with it the diffusion — that the catch and
yield could not. This is why `matchCatch()` matches all three: catch shape,
yield and production.

It is worth being honest about how much this buys you. Here production
changes by only about 20% across a four-fold change in mortality, because
the external mortality being tuned is a minor share of the total. So even
production constrains it only loosely: a modest error in the observed
production translates into a sizeable uncertainty in mortality. When the
external mortality is a small part of the budget like this, the most
reliable course is often to fix the diffusion (or the mortality) at an
externally informed value rather than to ask the fit to discover it.

## Does the optimiser find the true parameters?

So far we have probed the objective by hand, holding mortality fixed. A
separate, practical question is what the optimiser itself does when set
loose: does it recover the parameters that generated the data, and does the
answer depend on where it starts? We can check directly. We generate the
catch from the true model, then start `matchCatch()` from a range of
deliberately *wrong* values for the selectivity, mortality and diffusion,
and see where it lands — first with no production constraint, then with the
true production supplied.

```{r}
#| message: false
#| warning: false
g_true    <- gear_params(p0); gi0 <- which(g_true$species == species)
l50_true  <- g_true$l50[gi0]
prod_true <- getProduction(p0)[species]

# Fit from a given (very different) starting point, optionally supplying the
# observed production.
fit_from <- function(l50_0, mu_0, D_0, use_production) {
    pp <- p0
    gp <- gear_params(pp); gi <- which(gp$species == species)
    gp$l50[gi] <- l50_0; gp$l25[gi] <- l50_0 - 4; gp$catchability[gi] <- 1
    gear_params(pp) <- gp
    sp <- species_params(pp)
    sp$mu_mat[s] <- mu_0; sp$D_ext[s] <- D_0
    sp$production_observed[s] <- if (use_production) prod_true else NA
    pp@ext_diffusion[s, ] <- D_0 * pp@w^(sp$n[s] + 1)
    species_params(pp) <- sp
    fit <- suppressWarnings(matchCatch(
        pp, species = species, catch = catch_obs, yield_lambda = 1,
        production_lambda = if (use_production) 1 else 0,
        map = list(m = factor(NA))))
    gf <- gear_params(fit); gi <- which(gf$species == species)
    spf <- species_params(fit)[s, ]
    data.frame(l50_start = l50_0, mu_start = mu_0, D_start = D_0,
               l50 = round(gf$l50[gi], 1), mu_mat = round(spf$mu_mat, 3),
               D_ext = signif(spf$D_ext, 3),
               production = round(getProduction(fit)[species], 3))
}

# Four very different starting points.
starts <- list(c(30, 0.10, 3), c(40, 0.60, 20),
               c(55, 1.50, 0.1), c(70, 3.00, 0.01))

cat(sprintf("truth:  l50 = %.1f   mu_mat = %.2f   D_ext = %.1f   production = %.3f\n",
            l50_true, sps$mu_mat, sps$D_ext, prod_true))
```

**Without a production constraint:**

```{r}
#| message: false
#| warning: false
do.call(rbind, lapply(starts, function(x) fit_from(x[1], x[2], x[3], FALSE))) |>
    print(row.names = FALSE)
```

Every start converges to the *same* solution — the optimiser is robust to
its initial values; we cannot make it diverge by starting it badly. But
look at *which* solution. The selectivity is essentially recovered (`l50` ≈
36 from every start, near the true 35.4), yet the mortality and diffusion
are not: every run settles on `mu_mat` ≈ 0.73 and `D_ext` ≈ 15, whereas the
truth is 0.29 and 12.4, and the implied production overshoots the true value
by about a quarter. The optimiser reliably finds a point that reproduces the
catch and the yield — but it is a point on the degenerate valley, not the
truth. (This is genuine: it is not the optimiser stopping short, nor a
numerical artefact — the catch+yield objective really does take its lowest
value here rather than at the true parameters.)

**With the production supplied:**

```{r}
#| message: false
#| warning: false
do.call(rbind, lapply(starts, function(x) fit_from(x[1], x[2], x[3], TRUE))) |>
    print(row.names = FALSE)
```

The production constraint changes the destination. Now every start
converges — again robustly — to `mu_mat` ≈ 0.28, `D_ext` ≈ 12 and a
production near the true 0.31: close to the true values (not exact, because,
as we saw, production constrains the mortality only loosely). What changed
between the two tables is not the *reliability* of convergence but its
*destination*. Without production the optimiser converges dependably to the
wrong mortality and diffusion; with production it converges dependably to
approximately the truth. The danger, then, is not a fragile optimiser that
needs careful starting values — it is a confident one that returns a clean,
repeatable answer which, absent the production data, is simply the wrong
point on the valley.

## A new dataset gives a new answer

The previous section fixed the data and varied the starting point. Now we do
the reverse: fix a sensible starting point and vary the *data*, by drawing
fresh random catch samples (multinomial draws) from the same true model. If
the parameters were well determined, every sample would give essentially the
same fit. They do not — and *how* they differ tells us exactly what kind of
problem this is.

```{r}
#| message: false
#| warning: false
# True per-bin catch probabilities (1-cm bins) from the true model.
true_rate <- p0@initial_n[s, ] * getFMort(p0)[s, ] * p0@dw
agg <- aggregate(true_rate, list(length = floor(lengths)), sum)
names(agg)[2] <- "rate"
agg <- agg[agg$rate > 0 & (agg$length + 1) <= l_max, ]
true_probs <- agg$rate / sum(agg$rate)
true_bins  <- agg$length

# Fit one random multinomial catch sample, always from the same start.
fit_sample <- function(seed, n_total, use_production) {
    set.seed(seed)
    cnt <- as.vector(rmultinom(1, n_total, true_probs))
    catch <- data.frame(species = species, gear = "total",
                        length = true_bins[cnt > 0], dl = 1, catch = cnt[cnt > 0])
    pp <- p0
    gp <- gear_params(pp); gi <- which(gp$species == species)
    gp$l50[gi] <- 45; gp$l25[gi] <- 41           # same (mediocre) start every time
    gear_params(pp) <- gp
    if (use_production) {
        spp <- species_params(pp); spp$production_observed[s] <- prod_true
        species_params(pp) <- spp
    }
    fit <- suppressWarnings(matchCatch(
        pp, species = species, catch = catch, yield_lambda = 1,
        production_lambda = if (use_production) 1 else 0, map = list(m = factor(NA))))
    spf <- species_params(fit)[s, ]
    c(mu_mat = spf$mu_mat, D_ext = spf$D_ext)
}

seeds <- 1:6
big_noprod   <- t(sapply(seeds, fit_sample, n_total = 1e5,  use_production = FALSE))
small_noprod <- t(sapply(seeds, fit_sample, n_total = 2000, use_production = FALSE))
big_prod     <- t(sapply(seeds, fit_sample, n_total = 1e5,  use_production = TRUE))
```

```{r}
#| fig-cap: "Fitted (mu_mat, D_ext) for six random catch datasets drawn from the same true model (star). Without production, a large sample (filled blue) clusters tightly but away from the truth (bias), and a small sample (open grey) scatters widely along the valley (variance). With production (red triangles) every dataset lands near the truth."
plot(NULL, xlim = c(0, 1.5), ylim = c(9, 20),
     xlab = "fitted mu_mat", ylab = "fitted D_ext",
     main = "Six random datasets from one true model")
points(small_noprod[, 1], small_noprod[, 2], pch = 1,  col = "grey45", cex = 1.4, lwd = 2)
points(big_noprod[, 1],   big_noprod[, 2],   pch = 19, col = "blue")
points(big_prod[, 1],     big_prod[, 2],     pch = 17, col = "red")
points(sps$mu_mat, sps$D_ext, pch = 8, cex = 2.2, lwd = 2)
legend("topright", bty = "n",
       pch = c(8, 19, 1, 17), col = c("black", "blue", "grey45", "red"),
       pt.cex = c(1.5, 1, 1.4, 1),
       legend = c("truth", "no production, n = 1e5",
                  "no production, n = 2000", "with production, n = 1e5"))
```

Each fit still converges cleanly — the optimiser is no less reliable here
than before. But the answer it returns depends on the data, in two distinct
ways:

- **Variance.** With a smaller sample (open grey) the fits scatter widely
  along the valley — `mu_mat` ranges over more than a factor of two across
  the six datasets, with `D_ext` tracking it. This spread is sampling variability,
  and it is large *because* the likelihood is so flat: a little noise tilts a
  nearly level surface into a minimum somewhere new. It shrinks as the sample
  grows (the filled blue points, at fifty times the sample size, cluster much
  more tightly).
- **Bias.** But the blue cluster, tight as it is, sits at `mu_mat` ≈ 0.73,
  not at the truth (the star at 0.29). That offset is *not* sampling noise:
  it does not shrink with sample size, because the catch+yield objective is
  genuinely minimised away from the true value. More data buys you precision
  around the wrong answer.

Throughout, the selectivity and catchability (not plotted) barely move from
dataset to dataset — it is specifically the mortality–diffusion pair that is
unstable. And **the production constraint removes both problems at once**
(red triangles): every dataset lands tightly near the truth. Production
supplies the missing curvature, turning the flat valley into a genuine
basin, so the estimator becomes both low-variance and nearly unbiased.

## Takeaways

- The catch size distribution reflects the population spectrum $N(w)$
  filtered through selectivity. $N$ itself is shaped by growth, diffusion
  and mortality together, and **mortality and diffusion act on the
  upper tail in opposing ways** — more mortality removes large fish, more
  diffusion supplies them — so the catch shape cannot separate them.
- Fitting the catch shape well does **not** mean the parameters are
  correct. A range of (mortality, diffusion) pairs that rise and fall
  together reproduce the same catch.
- With **yield known but production unknown**, you can trust the
  selectivity and the catchability, but mortality and diffusion sit on a
  shallow valley along which catch and yield hardly change.
- Production responds to mortality and so helps close the valley, but its
  leverage depends on how large a share of total mortality the tuned
  external mortality represents — often small. Treat a yield-only
  calibration's mortality, diffusion and production as weakly determined,
  and supply an external constraint where you can.
- The optimiser is **robust to its starting values** — it returns the same
  fit from very different starts. A clean, repeatable result is therefore
  *not* evidence that the parameters are identified: without production the
  optimiser repeatably returns the wrong mortality and diffusion. Robustness
  of convergence and correctness of the answer are different things.
- A **new dataset gives a new answer**. Without production the fitted
  mortality and diffusion scatter from sample to sample (variance, growing as
  the sample shrinks) around a value that is itself off the truth (bias, not
  cured by more data). Production removes both, clustering every dataset near
  the true values.
```

