---
title: "How precisely does the catch size distribution determine the parameters?"
author: "Gustav Delius"
format:
  html:
    toc: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{How precisely does the catch size distribution determine the parameters?}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.2)
```

## The question

`matchCatch()` adjusts gear selectivity, catchability, mortality and the
growth-diffusion coefficient so that the model reproduces an observed catch
size distribution. A natural hope is that fitting the catch recovers the
*true* values of those parameters. How well does it?

The answer turns out to be uneven, and in an instructive way. Two of the
parameters — the gear selectivity and the catchability — are pinned down
sharply. The other two — mortality and diffusion — are recovered only
*loosely*: the fit is essentially unbiased (a perfect, noise-free catch
recovers the truth), but it is imprecise, because mortality and diffusion
trade off against each other along a shallow valley in which the catch and
yield barely change. At the sample sizes available in practice that
imprecision is large. The single most effective remedy is an independent
estimate of **production**, which we usually lack — so this vignette looks
in particular at the case that matters most: **the total yield is known but
the production is not.**

This is not a weakness of the optimiser; it is a property of how much
information the catch shape carries. The test
`"matchCatch recovers a model-generated catch size distribution"` checks
only that the *fitted catch distribution* matches the observed one, and its
comment warns that "the gear and mortality parameters themselves are not
uniquely recovered". This vignette builds intuition for *why*, through
examples.

## The catch is a product, but the population is not free

The model's predicted catch density at weight $w$ is

$$
C(w) \;=\; N(w)\, F(w),
\qquad F(w) = q\, S(w),
$$

where $N(w)$ is the steady-state number density, $S(w)$ the gear
selectivity (between 0 and 1) and $q$ the catchability. This is exactly
what the package computes — in `plot_catch()` the modelled catch is
`initial_n * getFMort`.

It would be glib to say "the catch is a product, so you cannot recover its
factors", because the population $N$ is **not** a free factor we may choose
at will. In the steady state, $N(w)$ is determined by the biology: it is
the solution of the size-structured balance between

- **growth** $g(w)$ — the deterministic rate at which an individual gains
  weight,
- **diffusion** $d(w) = D_{\text{ext}}\, w^{\,n+1}$ — the stochastic spread
  in growth, so that individuals of the same age reach a *range* of sizes,
- **mortality** $\mu(w)$ — which removes individuals and steepens the
  spectrum.

The real question is therefore sharper: given the catch shape, how well can
we disentangle *selectivity* from the *biological parameters that shape the
spectrum* — chiefly mortality and diffusion? Selectivity comes out cleanly.
Mortality and diffusion do not, and the reason is that **they both reshape
the upper tail of $N(w)$ in nearly the same way, so the catch shape responds
mostly to their combined effect and only weakly to the difference between
them.** That weak response is enough to recover them from a perfect catch,
but easily swamped by noise in a real one.

## A look at the model

Throughout we use a single-gear Celtic Sea model and generate catch data
from it, so that the truth is known and we can ask whether fitting the
catch recovers it. The helpers below build that model, sample its predicted
catch into 1-cm length bins, and let us set the diffusion coefficient.

```{r}
#| message: false
#| warning: false
library(mizerEcopath)

# Collapse celtic_params to one gear per species.
make_single_gear <- function() {
    params <- celtic_params
    gp <- gear_params(params)
    keep <- do.call(rbind, lapply(split(gp, gp$species), function(g) {
        row <- g[which.max(g$yield_observed), , drop = FALSE]
        row$gear <- "total"
        row$yield_observed <- sum(g$yield_observed)
        row
    }))
    rownames(keep) <- paste(keep$species, keep$gear, sep = ", ")
    gear_params(params) <- keep
    initial_effort(params) <- 1
    params
}

# Expected catch in each 1-cm length bin: the integral of the catch density
# N*F over the bin. We use the same piecewise-linear interpolation of the
# density that matchCatch()'s likelihood uses internally, so that the
# synthetic data is *self-consistent* with the model. (A cruder histogram —
# assigning each weight bin to floor(length) — disagrees with the likelihood's
# integration by a few percent, and because the mortality-diffusion valley is
# so flat that small discrepancy is enough to shift the fitted mortality
# substantially. Self-consistent binning is what lets a perfect catch recover
# the truth.)
model_catch_rate <- function(params, species, bins) {
    sp <- species_params(params); s <- which(sp$species == species); sps <- sp[s, ]
    w <- params@w
    dens <- approxfun(w, params@initial_n[s, ] * getFMort(params)[s, ], rule = 2)
    vapply(bins, function(L) {
        lo <- sps$a * L^sps$b; hi <- sps$a * (L + 1)^sps$b
        grid <- sort(unique(c(w[w > lo & w < hi], lo, hi)))
        v <- dens(grid); sum(diff(grid) * (head(v, -1) + tail(v, -1)) / 2)
    }, numeric(1))
}

# Generate model catch data over 1-cm bins (integer counts summing to ~n_total).
model_catch_data <- function(params, species, n_total = 1e5) {
    sp <- species_params(params); s <- sp$species == species; sps <- sp[s, ]
    w <- params@w; dens <- params@initial_n[s, ] * getFMort(params)[s, ]
    l_max <- (sps$w_max / sps$a)^(1 / sps$b)
    bins  <- seq(floor((min(w[dens > 0]) / sps$a)^(1 / sps$b)), floor(l_max) - 1)
    rate  <- model_catch_rate(params, species, bins)
    bins  <- bins[rate > 0]; rate <- rate[rate > 0]
    count <- round(rate / sum(rate) * n_total)
    data.frame(species = species, gear = "total", length = bins[count > 0],
               dl = 1, catch = count[count > 0])
}

# The model's predicted catch density (a PDF) over given length bins.
model_catch_pdf <- function(params, species, bins) {
    r <- model_catch_rate(params, species, bins)
    r / sum(r)
}

# Set the diffusion coefficient D_ext and recompute the (diffusion-aware)
# steady state, holding mortality, growth and selectivity fixed.
set_D_ext <- function(params, species, D) {
    sp <- species_params(params); s <- sp$species == species
    sp$D_ext[s] <- D; species_params(params) <- sp
    params@ext_diffusion[s, ] <- D * params@w^(sp$n[s] + 1)
    steadySingleSpecies(params, species = species)
}

species <- "Haddock"
# Solve the single-species steady state, the spectrum matchCatch() reconstructs,
# so the synthetic catch is one the fit's solver can reproduce exactly.
p0      <- steadySingleSpecies(make_single_gear(), species = species)
s       <- species_params(p0)$species == species
sps     <- species_params(p0)[s, ]
lengths <- (p0@w / sps$a)^(1 / sps$b)   # length (cm) of each weight bin

# Make the experiment self-consistent: the "observed" yield is this model's own
# yield, not the celtic multispecies value carried over by make_single_gear().
# The catch shape barely constrains the mortality, so an inconsistent yield
# target would pull the recovered mortality off the truth.
gp0 <- gear_params(p0); gi0 <- which(gp0$species == species)
gp0$yield_observed[gi0] <- getYield(p0)[s]
gear_params(p0) <- gp0
```

## How diffusion shapes the spectrum

Diffusion is the parameter readers tend to have least intuition for, so it
is worth pinning down with a measurement.

Without diffusion ($D_{\text{ext}} = 0$) every individual of a cohort
follows the same deterministic growth curve. As a fish matures it channels a
growing share of its surplus energy into reproduction, so its growth *slows
sharply* well short of the maximum size $w_{\max}$. For this Haddock model
growth has all but stalled by about 56 cm — only a few percent of its rate
at maturity — whereas $w_{\max}$ corresponds to 112 cm. With no diffusion,
then, individuals creep past 56 cm only very slowly while mortality keeps
removing them, so the spectrum *falls away sharply* there and the upper size
range is nearly empty.

Diffusion lets growth rates vary between individuals: a cohort *spreads
out* into a range of sizes as it ages (this is what the age-fitting
machinery in `project_diffusion()` exploits; see the *Diffusion*
vignette). Crucially, it carries some individuals *through* the near-stall
and on towards $w_{\max}$. The larger the diffusion, the more individuals
populate that upper tail. So **more diffusion means *more* large fish** — it
is diffusion, not deterministic growth, that fills the large-size end of the
spectrum.

We can see it directly in the steady-state spectra. Holding mortality and
selectivity fixed, we recompute the spectrum at three diffusion levels and
rescale each to the **same total biomass**, so the comparison is fair: any
difference is then in how that fixed biomass is *distributed* over sizes,
not in how much there is. We plot biomass density, so the area under each
curve is the (equal) total biomass. At low diffusion the biomass is bottled
up below ~56 cm, where growth stalls (a sharp drop there); raising the
diffusion spreads it out into the large-size tail:

```{r}
#| message: false
#| warning: false
#| fig-cap: "Steady-state biomass density at three diffusion coefficients, each rescaled to the same total biomass (equal area under the curve), with mortality and selectivity held fixed. At low `D_ext` the spectrum drops away sharply near ~56 cm, where growth stalls; larger `D_ext` fills the large-size tail. More diffusion means more large fish."
l_mat <- (sps$w_mat / sps$a)^(1 / sps$b)   # length at maturity (cm)
l_max <- (sps$w_max / sps$a)^(1 / sps$b)   # maximum length (cm)
dl   <- p0@dw / (sps$a * sps$b * lengths^(sps$b - 1))  # width of each bin (cm)
Ds   <- c(0.1, 10, 100)
cols <- c("blue", "black", "red")
rng  <- lengths >= 0.5 * l_mat & lengths <= 0.97 * l_max

# Biomass density in length, rescaled so the total biomass equals 1.
biomass_density <- function(D) {
    pp <- set_D_ext(p0, species, D)
    Bm <- pp@initial_n[s, ] * p0@w * p0@dw   # biomass in each bin
    (Bm / dl) / sum(Bm)
}
beta <- lapply(Ds, biomass_density)
ymax <- max(vapply(beta, function(b) max(b[rng]), numeric(1)))

plot(NULL, xlim = c(0.5 * l_mat, 0.97 * l_max), ylim = c(0, ymax),
     xlab = "Length [cm]", ylab = "Biomass density (total biomass = 1)",
     main = "Biomass spectrum at equal total biomass")
for (i in seq_along(Ds)) lines(lengths[rng], beta[[i]][rng], col = cols[i], lwd = 2.5)
abline(v = l_mat, lty = 3); text(l_mat, ymax, " maturity", adj = c(0, 1), cex = 0.8)
legend("topright", paste0("D_ext = ", Ds), col = cols, lwd = 2.5, bty = "n")
```

The effect on the catch follows directly. Raising `D_ext` lets the catch
reach larger sizes: with little diffusion almost nothing is caught above
60 cm, while at high diffusion a sixth of the catch is, and the mean catch
length climbs:

```{r}
#| message: false
#| warning: false
diff_demo <- do.call(rbind, lapply(c(1e-3, 0.1, 1, 10, 100), function(D) {
    pp <- set_D_ext(p0, species, D)
    cw <- pp@initial_n[s, ] * getFMort(pp)[s, ] * pp@dw
    cw <- cw / sum(cw)                       # normalised catch over weight
    data.frame(D_ext                = D,
               frac_catch_over_60cm = round(sum(cw[lengths > 60]), 3),
               mean_catch_length_cm = round(sum(cw * lengths), 1))
}))
print(diff_demo, row.names = FALSE)
```

Mortality and diffusion therefore push the large-fish abundance in
*opposite* directions: more mortality removes large fish, more diffusion
supplies them. That is the root of the trade-off. To keep the catch *almost*
unchanged while *raising* mortality (which removes large fish), the fit can
simply *raise* the diffusion (which puts them back), and vice versa. Because
the catch reveals mostly the net upper-tail shape (filtered through
selectivity), it constrains the two only loosely as a pair: a wide range of
mortality–diffusion combinations, rising and falling together, fit nearly
equally well. The combinations are not *exactly* equivalent — the catch does
faintly prefer the true pair — but the preference is slight, so it takes a
very clean catch to act on it.

## Yield known, production unknown

This is the situation we usually face: the total landed weight of a stock
is recorded, but an independent estimate of its production is not. What can
we then learn, and what stays out of reach?

We hold mortality (`mu_mat`) fixed at a range of values spanning a factor
of four around the truth, and at each one fit the catch shape **and** the
yield, but supply **no** production penalty (`production_lambda = 0`). The
gear selectivity, catchability and diffusion are free. If mortality were
identifiable from the catch and yield, only the true value would fit well.

```{r}
#| message: false
#| warning: false
catch_obs <- model_catch_data(p0, species)
obs_pdf   <- catch_obs$catch / sum(catch_obs$catch)
mu0       <- sps$mu_mat

mus <- c(0.5, 0.75, 1, 1.5, 2) * mu0
fits <- lapply(mus, function(mu) {
    pp <- p0
    sp <- species_params(pp); sp$mu_mat[s] <- mu; species_params(pp) <- sp
    suppressWarnings(matchCatch(
        pp, species = species, catch = catch_obs,
        yield_lambda = 1, production_lambda = 0,           # yield only
        map = list(m = factor(NA), mu_mat = factor(NA))))  # mortality fixed
})

summary_tab <- do.call(rbind, Map(function(mu, fit) {
    gpf <- gear_params(fit); gpf <- gpf[gpf$species == species, ]
    spf <- species_params(fit)[s, ]
    pm  <- model_catch_pdf(fit, species, catch_obs$length)
    tv  <- 0.5 * sum(abs(obs_pdf - pm))
    nll <- -sum(catch_obs$catch * log(pmax(pm, 1e-300)))  # catch-shape NLL
    data.frame(
        mu_mat       = round(mu, 3),
        l50          = round(gpf$l50, 1),
        catchability = round(gpf$catchability, 3),
        D_ext        = signif(spf$D_ext, 3),
        TV_distance  = round(tv, 4),
        NLL          = nll,
        yield        = signif(getYield(fit)[s], 3),
        production   = signif(getProduction(fit)[species], 3))
}, mus, fits))
# Show the catch-shape NLL relative to the best-fitting row; the sample size
# is the same in every row, so these differences are directly comparable.
summary_tab$NLL <- round(summary_tab$NLL - min(summary_tab$NLL), 1)
print(summary_tab, row.names = FALSE)
```

Three things stand out, and together they tell the whole story.

**The yield fixes the catchability.** Across the whole range the matched
yield sits on the observed value and the catchability barely moves. The
total landed mass $\int C(w)\,w\,dw = q\!\int S(w)\,N(w)\,w\,dw$ depends
directly on $q$, so requiring the model to reproduce the observed yield
determines $q$ — the one thing the *shape* alone leaves undetermined. This
is the part of the problem that yield data solve cleanly.

**The catch shape constrains the mortality only loosely.** Every row
reproduces the catch distribution closely — the total-variation distances
are all tiny — even though the assumed mortality ranges over a factor of
four. The fitted catch curves lie almost on top of one another:

```{r}
#| fig-cap: "Catch densities fitted at mortality assumptions spanning a factor of four, with the yield matched but no production constraint. The observed catch (grey bars) is reproduced almost equally well by all of them."
heights <- obs_pdf
plot(NULL, xlim = range(catch_obs$length), ylim = c(0, max(heights) * 1.05),
     xlab = "Length [cm]", ylab = "Relative catch density",
     main = paste(species, "— one catch, a range of mortalities"))
rect(catch_obs$length - 0.5, 0, catch_obs$length + 0.5, heights,
     col = "grey85", border = NA)
cols <- c("blue", "cyan4", "black", "orange", "red")
for (i in seq_along(fits)) {
    lines(catch_obs$length, model_catch_pdf(fits[[i]], species, catch_obs$length),
          col = cols[i], lwd = 2)
}
legend("topright", legend = paste0("mu_mat = ", round(mus, 2)),
       col = cols, lwd = 2, bty = "n")
```

The `TV_distance` column is the total-variation distance between the
observed and fitted catch — a readable measure of shape mismatch. The `NLL`
column is the catch-shape negative log-likelihood, the quantity
`matchCatch()` actually minimises, shown relative to the best-fitting row.
(Its absolute level is a large constant that every row carries; only the
differences matter, and since the sample size is the same in every row they
are directly comparable.) The likelihood tells a one-sided story that the
uniformly tiny total-variation distances hide. It has a shallow minimum
right at the true mortality (essentially tied with the value just below),
and from half the true value up to it the NLL varies by only a few units;
but above the truth it climbs by tens of units. So the catch pins the mortality mainly
*from above*: it clearly rules out values much larger than the truth, while
distinguishing the truth from lower values only feebly. That lower side is
what the yield and production must resolve. Note, though, that this is on a
likelihood scale: these "couple of units" are differences in the
log-likelihood of a 100,000-fish sample, so they are not negligible — they
are exactly what lets a clean catch recover the truth, as we will see.

**Diffusion is what absorbs the difference.** Look at the `D_ext` column:
as the assumed mortality *rises*, the fitted diffusion *rises* with it.
This is exactly the trade-off of the previous section. A higher mortality
removes large fish; the fit simply raises the diffusion to supply more, so
the upper tail of the catch is restored. The two move *together*, sliding
along a valley in which catch and yield stay almost constant — most flatly
on the low-mortality side, where even the likelihood barely separates the
fits.

Note how *mild* the diffusion adjustment is here — `D_ext` moves only from
about 11 to 14 across the whole four-fold range of mortality, not over
orders of magnitude. A large change in assumed mortality is offset by only a
small change in diffusion, and yet leaves the catch almost unchanged: the
catch is simply not very sensitive to *where*, along this trade-off, the
upper tail of the spectrum comes from. That insensitivity is the whole
difficulty in one sentence.

## The one quantity that does respond: production

The last column is the one that moves with mortality. The **production** —
the rate at which biomass is removed by mortality — climbs steadily across
the fits, while the yield stays pinned to its observed value:

```{r}
#| fig-cap: "With the yield matched but production unconstrained, the fitted yield (open circles) is flat across mortality assumptions, while the implied production (filled circles) rises. Yield carries no information about mortality; production carries some."
op <- par(mar = c(4, 4, 2, 4))
plot(summary_tab$mu_mat, summary_tab$production, type = "b", pch = 19,
     xlab = "assumed mu_mat", ylab = "production", main = "")
par(new = TRUE)
plot(summary_tab$mu_mat, summary_tab$yield, type = "b", pch = 1, lty = 2,
     axes = FALSE, xlab = "", ylab = "", ylim = c(0, max(summary_tab$yield) * 1.3))
axis(4); mtext("yield", side = 4, line = 2.5)
legend("topleft", c("production", "yield"), pch = c(19, 1), lty = c(1, 2),
       bty = "n")
par(op)
```

So if you calibrate to **yield only**, the result is:

| Quantity | Status after a yield-only fit |
|----------|-------------------------------|
| Gear selectivity ($l_{50}, l_{25}$) | **Well determined** by the catch shape |
| Catchability $q$ | **Well determined** by the yield |
| Mortality (`mu_mat`) | **Unbiased but imprecise** — recovered on average, bounded from above by the catch, loose below; trades off against diffusion |
| Diffusion (`D_ext`) | **Unbiased but imprecise** — trades off against mortality; precision falls as the sample shrinks |
| Production | Follows the imprecise mortality |

The model you get back will reproduce the observed catch and yield
faithfully, and its selectivity and catchability can be trusted. Its
mortality and diffusion are correct *on average* but sit on a shallow valley
along which the catch and yield hardly change, so any single fit can be far
off — by a factor of two or more at realistic sample sizes, as we will see.
Production moves along that valley, so it is the natural quantity to bring in
if you want to sharpen the mortality.

## What closes the gap

Adding an observed production to the objective (`production_lambda > 0`)
gives the optimiser a constraint that *does* vary along the valley, so it
sharpens the mortality — and with it the diffusion. The catch and yield pin
the mortality only weakly, through the faint downhill tilt of the valley
floor: that tilt is enough on noise-free data (as the next section shows) but
little defence once sampling noise enters, whereas production grips the
mortality directly. This is why `matchCatch()` matches all three: catch
shape, yield and production.

It is worth being honest about how much this buys you. An *accurate*
production pins the mortality far more tightly — we will see the per-dataset
scatter shrink severalfold. But the leverage is limited: here production
changes by only about 20% across a four-fold change in mortality, so the
inference runs the other way too — a modest *error* in the observed
production maps into a sizeable error in the inferred mortality. When even
production is uncertain, the most reliable course is often to fix the
diffusion (or the mortality) at an externally informed value rather than to
ask the fit to discover it.

## Does the optimiser find the true parameters?

So far we have probed the objective by hand, holding mortality fixed. A
separate, practical question is what the optimiser itself does when set
loose: does it recover the parameters that generated the data, and does the
answer depend on where it starts? We use here the model's *exact* predicted
catch — the noise-free expected counts of `model_catch_data()`, with no
sampling scatter — and start `matchCatch()` from a range of deliberately
*wrong* values for the selectivity, mortality and diffusion, to see where it
lands; first with no production constraint, then with the true production
supplied.

```{r}
#| message: false
#| warning: false
g_true    <- gear_params(p0); gi0 <- which(g_true$species == species)
l50_true  <- g_true$l50[gi0]
prod_true <- getProduction(p0)[species]

# Fit from a given (very different) starting point, optionally supplying the
# observed production.
fit_from <- function(l50_0, mu_0, D_0, use_production) {
    pp <- p0
    gp <- gear_params(pp); gi <- which(gp$species == species)
    gp$l50[gi] <- l50_0; gp$l25[gi] <- l50_0 - 4; gp$catchability[gi] <- 1
    gear_params(pp) <- gp
    sp <- species_params(pp)
    sp$mu_mat[s] <- mu_0; sp$D_ext[s] <- D_0
    sp$production_observed[s] <- if (use_production) prod_true else NA
    pp@ext_diffusion[s, ] <- D_0 * pp@w^(sp$n[s] + 1)
    species_params(pp) <- sp
    fit <- suppressWarnings(matchCatch(
        pp, species = species, catch = catch_obs, yield_lambda = 1,
        production_lambda = if (use_production) 1 else 0,
        map = list(m = factor(NA))))
    gf <- gear_params(fit); gi <- which(gf$species == species)
    spf <- species_params(fit)[s, ]
    data.frame(l50_start = l50_0, mu_start = mu_0, D_start = D_0,
               l50 = round(gf$l50[gi], 1), mu_mat = round(spf$mu_mat, 3),
               D_ext = signif(spf$D_ext, 3),
               production = round(getProduction(fit)[species], 3))
}

# Four very different starting points.
starts <- list(c(30, 0.10, 3), c(40, 0.60, 20),
               c(55, 1.50, 0.1), c(70, 3.00, 0.01))

cat(sprintf("truth:  l50 = %.1f   mu_mat = %.2f   D_ext = %.1f   production = %.3f\n",
            l50_true, sps$mu_mat, sps$D_ext, prod_true))
```

**Without a production constraint:**

```{r}
#| message: false
#| warning: false
do.call(rbind, lapply(starts, function(x) fit_from(x[1], x[2], x[3], FALSE))) |>
    print(row.names = FALSE)
```

Every start converges to the *same* solution — the optimiser is robust to
its initial values; we cannot make it diverge by starting it badly. And the
solution is essentially the truth: `l50` ≈ 35.4, `mu_mat` ≈ 0.29 and
`D_ext` ≈ 12.4, against the true 35.4, 0.29 and 12.4. So from a *perfect*
catch the parameters can be recovered — even the mortality and diffusion,
despite the shallow valley between them, and *without* the production
constraint. The catch is not fundamentally ambiguous: the valley floor
tilts, ever so slightly, toward the truth, and with noise-free data that
faint tilt is enough.

**With the production supplied:**

```{r}
#| message: false
#| warning: false
do.call(rbind, lapply(starts, function(x) fit_from(x[1], x[2], x[3], TRUE))) |>
    print(row.names = FALSE)
```

Supplying the production changes the destination hardly at all here —
`mu_mat` ≈ 0.29, `D_ext` ≈ 12.4, production ≈ 0.32 — because with a noise-free
catch the optimiser already had what it needed to find the truth. On perfect
data the two calibrations agree, and both are essentially right. That is the
reassuring part, and it is worth stating plainly: the objective is
well-posed and its minimum sits at the truth, reached robustly from any
starting point. The parameters are *identifiable in principle*.

The catch (with the yield) really does, then, contain the information needed
to recover the mortality and diffusion — but only a little of it, concentrated
in that faint downhill tilt of the valley floor. Real data are not noise-free,
and the next section shows what happens to that small signal once sampling
noise is added: it is where production earns its keep.

## A new dataset gives a new answer

The previous section fixed the data and varied the starting point. Now we do
the reverse: fix a sensible starting point and vary the *data*, by drawing
fresh random catch samples (multinomial draws) from the same true model. If
the parameters were well determined, every sample would give essentially the
same fit. They do not — and *how* they differ tells us exactly what kind of
problem this is.

```{r}
#| message: false
#| warning: false
# The model's exact expected catch per 1-cm bin (same binning as the
# likelihood), used as the multinomial probabilities to draw samples from.
true_bins  <- catch_obs$length
true_probs <- model_catch_pdf(p0, species, true_bins)

# Fit one random multinomial catch sample, always from the same start.
fit_sample <- function(seed, n_total, use_production) {
    set.seed(seed)
    cnt <- as.vector(rmultinom(1, n_total, true_probs))
    catch <- data.frame(species = species, gear = "total",
                        length = true_bins[cnt > 0], dl = 1, catch = cnt[cnt > 0])
    pp <- p0
    gp <- gear_params(pp); gi <- which(gp$species == species)
    gp$l50[gi] <- 45; gp$l25[gi] <- 41           # same (mediocre) start every time
    gear_params(pp) <- gp
    if (use_production) {
        spp <- species_params(pp); spp$production_observed[s] <- prod_true
        species_params(pp) <- spp
    }
    fit <- suppressWarnings(matchCatch(
        pp, species = species, catch = catch, yield_lambda = 1,
        production_lambda = if (use_production) 1 else 0, map = list(m = factor(NA))))
    spf <- species_params(fit)[s, ]
    c(mu_mat = spf$mu_mat, D_ext = spf$D_ext)
}

# A realistic sample size (n = 2000), with and without production, plus a very
# large sample without production to show that the spread is variance (it
# shrinks with more data), not bias.
seeds       <- 1:6
noprod_2000 <- t(sapply(seeds, fit_sample, n_total = 2000, use_production = FALSE))
prod_2000   <- t(sapply(seeds, fit_sample, n_total = 2000, use_production = TRUE))
noprod_1e5  <- t(sapply(seeds, fit_sample, n_total = 1e5,  use_production = FALSE))
```

```{r}
#| fig-cap: "Fitted (mu_mat, D_ext) for six random catch datasets of a realistic size (n = 2000 fish) drawn from the same true model (star). Without production (open circles) the fits scatter widely along the mortality-diffusion valley but centre on the truth. With production (red triangles) the mortality is pinned on it (the diffusion stays looser, spreading vertically). The scatter is variance, not bias."
xr <- range(c(noprod_2000[, 1], prod_2000[, 1], sps$mu_mat))
yr <- range(c(noprod_2000[, 2], prod_2000[, 2], sps$D_ext))
plot(NULL, xlim = xr, ylim = yr,
     xlab = "fitted mu_mat", ylab = "fitted D_ext",
     main = "Six random datasets (n = 2000) from one true model")
points(noprod_2000[, 1], noprod_2000[, 2], pch = 1,  col = "blue", cex = 1.4, lwd = 2)
points(prod_2000[, 1],   prod_2000[, 2],   pch = 17, col = "red")
points(sps$mu_mat, sps$D_ext, pch = 8, cex = 2.2, lwd = 2)
legend("topleft", bty = "n",
       pch = c(8, 1, 17), col = c("black", "blue", "red"),
       pt.cex = c(1.5, 1.4, 1),
       legend = c("truth", "no production", "with production"))
```

Every fit still converges cleanly — the optimiser is reliable here as
before. But without production the answer it returns swings with the data:

- **It is variance, not bias.** Across the six datasets the no-production
  fits (open circles) scatter along the valley — `mu_mat` ranges over a
  factor of two or more, with `D_ext` tracking it — but they *centre on the
  truth* (the star). There is no systematic offset; a single fit is simply an
  imprecise draw. The spread is sampling variability, large *because* the
  likelihood is so flat that a little noise tilts a nearly level surface into
  a minimum somewhere new.
- **The spread shrinks with more data.** At fifty times the sample
  (`noprod_1e5`, not plotted) the no-production `mu_mat` tightens to roughly
  0.26–0.32 — still scattered, still centred on the truth, but far narrower.
  Precision, not correctness, is what the sample size buys; that is the
  signature of variance.

Throughout, the selectivity and catchability (not plotted) stay put from
dataset to dataset — even at `n = 2000`, `l50` holds to within a few tenths
of a centimetre. It is specifically the mortality–diffusion pair that is
unstable. And **the production constraint tames it** (red triangles): with an
accurate production the `mu_mat` estimates collapse onto a narrow band around
the truth (here roughly 0.25–0.29, several times tighter than without), even
as the diffusion stays looser — the red points still spread vertically —
because production bears directly on mortality, not on the diffusion, which
the catch tail alone must fix. Production supplies the steep constraint the
flat valley lacks along the mortality axis, turning an imprecise mortality
estimate into a much sharper one.

## Takeaways

- The catch size distribution reflects the population spectrum $N(w)$
  filtered through selectivity. $N$ itself is shaped by growth, diffusion
  and mortality together, and **mortality and diffusion act on the upper
  tail in nearly the same way** — more mortality removes large fish, more
  diffusion supplies them — so the catch responds strongly to their combined
  effect and only weakly to the difference between them.
- **Selectivity and catchability are well determined** — the first by the
  catch shape, the second by the yield — at every sample size we tried.
- **Mortality and diffusion are determined, but only loosely.** The fit is
  essentially unbiased — a perfect, noise-free catch recovers the truth, and
  the optimiser reaches it robustly from any starting point — so the
  parameters are identifiable in principle. But the valley between mortality
  and diffusion is so flat that realistic sampling noise scatters any single
  fit widely (a factor of two or more at `n ≈ 2000`). The trouble is
  precision, not bias, and not the optimiser.
- **A new dataset gives a new answer**, but a *centred* one: the no-production
  fits scatter around the truth and tighten as the sample grows. That is the
  hallmark of variance — fixed by more or better data, not by a better
  algorithm.
- **Production is the effective remedy.** An accurate production supplies the
  steep constraint the catch and yield lack under noise, pinning the mortality
  several times more tightly. Its leverage is limited, though — production
  changes only ~20% across a four-fold change in mortality — so an
  *inaccurate* production feeds error back into the mortality. When even production is uncertain,
  fixing the diffusion (or mortality) at an externally informed value is
  often the most reliable course.
```

