---
title: "Celtic_Sea_Status_Quo_Model"
author: "Jessica Westworth"
format: pdf
editor: visual
---

# Introduction

Creating a model representing the average state of the Celtic Sea from 2012 to 2024. Composed of 15 species: herring (*Clupea harengus*), cod (*Gadus morhua*), haddock (*Melanogrammus aeglefinus*), whiting (*Merlangius merlangus*), blue whiting (*Micromesistius poutassou*), hake (*Merluccius merluccius*), monkfish (Lophius piscatorius), horse mackerel (*Trachurus trachurus*), mackerel (*Scomber scombrus*), plaice (*Pleuronectes platessa*), megrim (*Lepidorhombus whiffiagonis*), sole (*Solea solea*), boarfish (*Capros aper*), dragonet (Callionymus spp.), and red gurnard (Chelidonicthys cuculus).

## Loading in Repository and Libraries

```{r}
#| warning: false
#| message: false
#| results: 'hide'

#library(remotes)
#install_github("sizespectrum/mizer", ref="diffusion") 
#install_github("gustavdelius/mizerEcopath", ref="spawning_season") 
library(mizer)
library(mizerEcopath)
library(dplyr)
library(here)
```

## Load in Data

Data includes: life_history parameters (l_mat and age_mat calculated from DATRAS data; w_max and LWR from Fishbase); landing distributions form STECF; survey size distribution and age at length distribution data from DATRAS; Stomach contents data; Biomass estimates, yield estimates, interaction matrix, and diet matrix from Lauria (2016).

```{r}
#| warning: false
#| message: false
#| results: 'hide'

#Life history data

#Load in ecopath estimates

    #Biomass

    #Yield

    #Interaction matrix

    #Diet matrix

#Landings size distibutions data

#DATRAS size distribution

#DATRAS age at length distribution

```

# Step 1

## Species Parameters

Assign species parameters from life history data

```{r}
# Set species and their scientific names
species <- data.frame(
    species = c("Herring", "Cod", "Haddock", "Whiting", "Blue whiting", "Hake", 
                "Monkfish", "Horse mackerel", "Mackerel", "Plaice", "Megrim", "Sole",
                "Boarfish" , "Dragonets" , "Red Gurnard"),
    
    Scientific_name = c("Clupea harengus", "Gadus morhua", "Melanogrammus aeglefinus", 
                        "Merlangius merlangus", "Micromesistius poutassou", 
                        "Merluccius merluccius", "Lophius piscatorius", 
                        "Trachurus trachurus", "Scomber scombrus", 
                        "Pleuronectes platessa", "Lepidorhombus whiffiagonis", 
                        "Solea solea", "Capros aper", "Callionymus spp.", 
                        "Chelidonicthys cuculus")
)

#Add life-history parameters corresponding to species

```
