---
title: "Celtic Sea Model"
format: html
editor: visual
---

# Introduction

Creating a Celtic Sea model including species herring (*Clupea harengus*), cod (*Gadus morhua*), haddock (*Melanogrammus aeglefinus*), whiting (*Merlangius merlangus*), blue whiting (*Micromesistius poutassou*), hake (*Merluccius merluccius*), monkfish (Lophius piscatorius), horse mackerel (*Trachurus trachurus*), mackerel (*Scomber scombrus*), plaice (*Pleuronectes platessa*), megrim (*Lepidorhombus whiffiagonis*), sole (*Solea solea*).

## Loading in Repository and Libraries

```{r}
#| warning: false
#| message: false
#| results: 'hide'
library(remotes)
#install_github("gustavdelius/mizerEcopath") 
library(mizerEcopath)
library(dplyr)
library(here)
```

## Load in Data

Load in the accompanying csv and rds files within the mizerEcopath repostiory into R objects. This data includes species parameters data diet data and survey fishing data taken from ecopath and DATRAS.

Some of the data formatting alterations are made in the following code to ensure that they are compatible with the desired MizerEcopath functions.

```{r}
#| warning: false
#| message: false
#| results: 'hide'

#Load in ecopath parameters Consumption 
ecopath_params <- read.csv(here("inst","extdata","Ecopath-Basic estimates.csv"))

#Ecopath yield data
yield <- read.csv(here("inst","extdata","Ecopath-Fishing mortality rates.csv"))
colnames(yield)[2] <- "Group.name"
yield$Total<-rowSums(yield[,3:29],na.rm=TRUE)
yield <- left_join(ecopath_params, yield, by = c("Group.name" = "Group.name")) |>
    mutate(TotalCatch..t.km..year. = Total * Biomass..t.km.2.) |>
    select(Group.name, TotalCatch..t.km..year.)

#Scaled landings data
cs_scaled_landings<- read.csv(here("inst","extdata", "cs_scaled_landings.csv")) 
cs_scaled_landings <- cs_scaled_landings[cs_scaled_landings$gear == "total", ] #only total gear
cs_scaled_landings <- cs_scaled_landings %>%
  mutate(length = floor(length)) %>%           # Round up lengths
  group_by(species, gear, length, Scientific_name) %>%  # Group by relevant columns
  summarise(catch = sum(scaled_catch, na.rm = TRUE),
            biomass = sum(scaled_biomass, na.rm = TRUE), .groups = "drop")                                                          # Sum catch for duplicates
cs_scaled_landings$count <- cs_scaled_landings$catch
cs_scaled_landings$dl<-1
cs_scaled_landings<-cs_scaled_landings[,c("species","Scientific_name","length","dl", "gear","catch","count", "biomass")]

#Survey data
cs_survey<- read.csv(here("inst","extdata", "cs_survey.csv")) 

cs_survey <- cs_survey %>%
  group_by(species, Scientific_name, length, dl) %>%
  summarise(
    gear = "total",                          # Create a new 'gear' category
    catch =sum(scaled_catch, na.rm = TRUE),       # Sum all catch across gears
    count= catch,
    biomass = sum(scaled_biomass, na.rm = TRUE),       # Sum all catch across gears
    .groups = "drop"
  )

#For the time being cs_survey is going to be called catch but in future we will combine the landings and survey and have them as different gear types.
catch<-cs_survey

#Future addition: combine survey and landings as different gear types catch? call this dataframe catch.

cs_survey$gear<-"survey"
cs_scaled_landings$gear<-"landings"
catch<-rbind(cs_survey,cs_scaled_landings)

#Age Length Data
age_at_length <- readRDS(here("inst", "extdata", "Celtic_Sea_Size_at_Age_Data.rds"))

#Species parameters from Fishbase
    #Fishbase growth parameters, growth parameters are chosen by list of studies     provided by fish base, study from which species growth parameters are taken were     selected by availability of data, and location proximity to the Celtic Sea. a     and b values are taken from the Bayesian length-weight ratio model estimated tab     in fishbase for each species.
life_histories <- read.csv(here("inst","extdata", "life_histories.csv"))

life_histories_summary <- life_histories %>%
  filter(Sex %in% c("Unsexed", "Mixed")) %>%      # keep Unsexed or Mixed
  mutate(Scientific_name = Species) %>%           # duplicate Species to match maturity data
  group_by(Species) %>%
  # Pick row with l_mat if possible, else any row
  slice(
    if (any(!is.na(l_mat))) {
      which(!is.na(l_mat))[1]  # first non-NA
    } else {
      1                        # first row if all NA
    }
  ) %>%
  rename(l_mat1 = l_mat) %>%
  mutate(
    Scientific_name = Species,
    l_max = case_when(
      l_max_type == "TL" ~ l_max_fishbase,
      l_max_type == "FL" ~ La + Lb * l_max_fishbase,
      l_max_type == "SL" ~ La + Lb * l_max_fishbase
    ),
    l_max = ifelse(!is.na(l_max), l_max, l_max_fishbase)
  ) %>%
  ungroup()

#length maturity from processed size at age DATRAS data
lm<-read.csv(here("inst","extdata", "cs_length_maturity_from_age_size_data.csv"))
lm<-rename(lm,l_mat2=l50)
sp_params<-left_join(life_histories_summary, lm, by="Scientific_name")

#assume life_histories summary l_mat1, lm l_mat2, l_mat is the new column
sp_params <- sp_params %>%
  mutate(
    l_mat = ifelse(
      !is.na(l_mat2),
      l_mat2,
      l_mat1
    ))%>%
 select(-l_mat2)%>%
 select(-l_mat1)
```

# Step 1

## 1.1 Species Parameters

```{r}
# Mapping of all species to their scientific names
species_map_lauria <- data.frame(
    species = c("Herring", "Cod", "Haddock", "Whiting", "Blue whiting", "Hake", 
                "Monkfish", "Horse mackerel", "Mackerel", "Plaice", "Megrim", "Sole"),
    Scientific_name = c("Clupea harengus", "Gadus morhua", "Melanogrammus aeglefinus", 
                        "Merlangius merlangus", "Micromesistius poutassou", "Merluccius merluccius",
                        "Lophius piscatorius", "Trachurus trachurus", "Scomber scombrus", 
                        "Pleuronectes platessa", "Lepidorhombus whiffiagonis", "Solea solea")
)

sp <- sp_params %>%
    # add computed columns first
    mutate(
        SciName = Species,                # keep scientific name
        t0      = ifelse(is.na(t0), 0, t0),
        l_mat=l_mat,
        l_max=l_max,
        w_mat   = a * (l_mat ^ b),       # weight at maturity
        w_max   = a * (l_max ^ b),       # max weight
        n       = 0.7,
        p       = 0.7,
        d       = n - 1,
        alpha   = 0.8
    ) %>%
        # join common names
    left_join(
        species_map_lauria,
        by = c("SciName" = "Scientific_name")
    ) %>%
    # overwrite species with common name
    mutate(species = species) %>%
    # keep desired columns
    transmute(
        species,     # common name 
        SciName,     # scientific name 
        w_inf,
        k_vb,
        l_mat,
        l_max,
        w_max,
        w_mat,
        a, b, t0,
        n, p, d, alpha
    )
sp$age_mat <- mizer:::age_mat_vB(sp)
```

## Add Ecopath Params

```{r}
## 4.2 Map to Ecopath stanzas
species_to_groups <- list(
    Herring       = "Herring",
    Cod           = c("Cod", "Juvenile cod"),
    Haddock       = c("Haddock","Juvenile haddock"),
    Whiting       = c("Whiting","Juvenile whiting"),
    `Blue whiting`= c("Blue whiting","Juvenile blue whiting"),
    Hake          = c("Hake","Juvenile hake"),
    Monkfish      = c("Monkfish","Juvenile monkfish"),
    `Horse mackerel` = "Horse mackerel",
    Mackerel      = "Mackerel",
    Plaice        = c("Plaice","Juvenile plaice"),
    Megrim        = c("Megrim","Juvenile megrim"),
    Sole          = "Sole"
)

# Use addEcopathParams to validate and add Ecopath parameters to Mizer species.
sp <- addEcopathParams(sp, ecopath_params, species_to_groups)
```

## Update w_max from observations

```{r}
## 5.3 Update w_max from observed sizes
observed_max <- cs_scaled_landings %>%
  group_by(species) %>%
  summarise(
    obs_len = if (all(is.na(length + dl))) NA_real_ else max(length + dl, na.rm = TRUE)
  )

sp1 <- sp %>%
    left_join(observed_max, by = "species")%>%
    mutate(
        l_max      = pmax(obs_len, l_max),
        w_max        = a * l_max^b,
    )
```

## Build non-interacting mizer model

```{r}
## 6.1 Initial model ----
model_non_int <- newAllometricParams(sp, no_w = 200)

## 6.2 Add survey ---
model_non_int <- addCatch(model_non_int, landings= cs_scaled_landings, survey=cs_survey,step=1)

# Match growth and biomass
# We need species_params to get the adult biomass
species_params <- species_params(model_non_int)

# Populate observed adult biomasses from Ecopath
for (i in seq_len(nrow(species_params))) {
    species_name <- species_params$species[i]
    stanza <- species_to_groups[[species_name]][1]  # leading (adult) stanza
    species_params$biomass_observed[i] <- ecopath_params %>%
        filter(Group.name == stanza) %>%
        pull(Biomass..t.km.2.) %>%
        sum(na.rm = TRUE)
}

# Set biomass_cutoff to maturation weight (w_mat)
species_params$biomass_cutoff <- species_params$w_mat

# Place biomass cutoff back into the species parameters
species_params(model_non_int) <- species_params

model_non_int <- model_non_int %>%
    matchGrowth() %>%
    mizerEcopath::steadySingleSpecies() %>%
    matchBiomasses()

model_non_int <- model_non_int %>%
    matchConsumption()

model_non_int_1<-tuneEcopath(
    model_non_int, 
    catch = cs_survey,
    age_at_length = age_at_length
)
```

# Step 2 - Adding Landings

```{r}
model_non_int <- addFishingMort(model_non_int, cs_scaled_landings)
```

# Step 3 - Diet Matching

```{r eval=FALSE}
fits <- readRDS(here("Data", "stomach_data_fit.rds"))
ecopath_diet <- read.csv(here("Data", "Ecopath-Diet composition.csv")) 
```
