---
title: "Celtic_Sea_Status_Quo_Model"
author: "Jessica Westworth"
format: pdf
editor: visual
---

# Introduction

Creating a model representing the average state of the Celtic Sea from 2012 to 2024. Composed of 15 species: herring (*Clupea harengus*), cod (*Gadus morhua*), haddock (*Melanogrammus aeglefinus*), whiting (*Merlangius merlangus*), blue whiting (*Micromesistius poutassou*), hake (*Merluccius merluccius*), monkfish (Lophius piscatorius), horse mackerel (*Trachurus trachurus*), mackerel (*Scomber scombrus*), plaice (*Pleuronectes platessa*), megrim (*Lepidorhombus whiffiagonis*), sole (*Solea solea*), boarfish (*Capros aper*), dragonet (Callionymus spp.), and red gurnard (Chelidonicthys cuculus).

## Loading in Repository and Libraries

```{r}
#| warning: false
#| message: false
#| results: 'hide'

#library(remotes)
#install_github("sizespectrum/mizer", ref="diffusion") 
#install_github("gustavdelius/mizerEcopath", ref="spawning_season") 
library(mizer)
library(mizerEcopath)
library(dplyr)
library(here)
```

## Load in Data

Data includes: life_history parameters (l_mat and age_mat calculated from DATRAS data; w_max and LWR from Fishbase); landing distributions form STECF; survey size distribution and age at length distribution data from DATRAS; Stomach contents data; Biomass estimates, yield estimates, interaction matrix, and diet matrix from Lauria (2016).

```{r}
#| warning: false
#| message: false
#| results: 'hide'

#Life history data from fishbase
data(life_history_fishbase)
life_history_fishbase$l_mat<-c(27.15, 42.36707, 23.473, 21.13458, 25.1, 32.87942, 55.0, 22.9, 28.06812, 17.73243, 20.62504, 20.8603,  9.2, 21.31807)

#maturity values calculated from survey data
#data(cs_calculated_histories)

    #combine life_history_fishbase and calculated l_mat
#life_histories<-life_history_fishbase%>%
#    mutate(
#        l_mat=case_when(
#        l_mat.y %in% c(NA_character_) ~ l_mat.x,
#        TRUE ~ l_mat.y),
#        age_mat=case_when(
#        age_mat.x %in% c(NA_character_) ~ age_mat.y,    
#        TRUE ~ age_mat.x))%>%
#    arrange(species)%>%
#    select(-l_mat.x, -l_mat.y, -age_mat.x, -age_mat.y)

#Load in ecopath data
ecopath_estimates <- read.csv(here("inst","extdata","Ecopath-Basic estimates.csv"))

Gurnards <- ecopath_estimates[46, ]
Boarfish <- ecopath_estimates[49, ]
num_cols <- sapply(Gurnards, is.numeric)
Red_gurnard<-Gurnards
Other_Gurnards<-Gurnards
Boarfish<-Boarfish
Small_pelag<-Boarfish
Red_gurnard[num_cols] <- Gurnards[num_cols] * 0.37
Red_gurnard$Group.name<-"Red gurnard"
Other_Gurnards[num_cols] <- Gurnards[num_cols] * 0.63
Boarfish[num_cols]<- Boarfish[num_cols] * 0.97
Boarfish$Group.name<-"Boarfish"
Small_pelag[num_cols]<- Boarfish[num_cols] * 0.03
#doesn't matter that tropic level and hab area are being affected as they are not used in calculations
ecopath_params<-ecopath_estimates[-c(46,49),]
ecopath_params<-rbind(ecopath_params,Other_Gurnards, Red_gurnard, Boarfish, Small_pelag)

#Diet matrix
data(diet)

#Deaths from fishing
data(fishing_deaths)

#Catch Distribution
data(catch_distribution)
catch_distribution<-catch_distribution%>%filter(!species=="Dragonet")


#DATRAS size distribution
data(survey_length_distribution)
survey_length_distribution<-survey_length_distribution%>%
    filter(!species=="Dragonet")

#DATRAS age at length distribution
data(cs_age_size)
boar<-readRDS("/Users/jessicawestworth/Desktop/boarfish_struct_like_DATRAS.rds")
cs_age_size<-rbind(cs_age_size,boar)

#determining l_max comparing landings max length versus fishbase max length recorded value
observed_max <- catch_distribution %>%
     group_by(species) %>%
     summarise(
         obs_len = if (all(is.na(length + dl))) NA_real_ else max(length + dl, na.rm = TRUE)
     )
life_histories <- life_history_fishbase %>%
    left_join(observed_max, by = "species")%>%
    mutate(
        l_max = pmax(obs_len, l_max))%>%
    select(-obs_len)

#combined survey fishing and STECF fishing
all_fishing<-rbind(catch_distribution,survey_length_distribution)
```

## Species Parameters

Assign species parameters from life history data

```{r}
#calculate w_mat and w_max and set n,p,d, and alpha parameter
sp_params<-life_histories%>%
    mutate(w_mat=a*l_mat^b,
           w_max=a*l_max^b,
           n=0.7,
           p=0.7,
           d=-0.3,
           alpha=0.7,
           biomass_cutoff=w_mat)

species_to_groups <- list(
    Herring          = "Herring",
    Cod              = c("Cod", "Juvenile cod"),
    Haddock          = c("Haddock","Juvenile haddock"),
    Whiting          = c("Whiting","Juvenile whiting"),
    `Blue whiting`   = c("Blue whiting","Juvenile blue whiting"),
    Hake             = c("Hake","Juvenile hake"),
    Monkfish         = c("Monkfish","Juvenile monkfish"),
    `Horse mackerel` = "Horse mackerel",
    Mackerel         = "Mackerel",
    Plaice           = c("Plaice","Juvenile plaice"),
    Megrim           = c("Megrim","Juvenile megrim"),
    Sole             = "Sole",
    Boarfish         = "Boarfish",
    `Red gurnard`      = "Red gurnard"
)

sp <- sp_params
sp <- addEcopathParams(sp, ecopath_params, species_to_groups)
```

## Build Non-Interacting Model

```{r}
model_non_int <- newAllometricParams(sp_params, no_w = 200)

## Add gear params
model_non_int <- addCatch(model_non_int, landings= catch_distribution,
                              fishing_dead_biomass=fishing_deaths,
                              survey= survey_length_distribution,step=2)

# Steady model
model_non_int <- model_non_int %>%
    steadySingleSpecies() %>%
    matchBiomasses() %>%
    steadySingleSpecies() %>%
    matchBiomasses()%>%
    steadySingleSpecies() %>%
    matchBiomasses()%>%
    steadySingleSpecies() %>%
    matchBiomasses()%>%
    steadySingleSpecies() %>%
    matchBiomasses()

params<-model_non_int
sps<-species_params(model_non_int)
initial_Eiw<-array(dim=length(sps$species))
for(i in 1:length(sps$species)){
    initial_Eiw[i]<-getExtEncounter(params)[i,1]/(w(params)[1]^sp[i,]$n)}
sps$Eiw<-initial_Eiw
species_params(model_non_int)<-sps

#set the spawning time, spawning concentration and mu mortality of species to predetermined values.

#spawning_mort<-tuneEcopath(
#    model_non_int, 
#    catch = all_fishing,
#    age_at_length = cs_age_size)

#saveRDS(spawning_mort,"/Users/jessicawestworth/Desktop/spawning_mort.rds")
```

Add diets and predation kernel

```{r}
#Adjust diet matrix for integration into model using reduceEcopathDiet
dm<-reduceEcopathDiet(
    species_params = species_params(spawning_mort),
    ecopath_diet   = diet,
    ecopath_params = ecopath_params
)

# Stomach-fit parameters (keep separate; align by scientific name), cannot do right now because the new species are not in the processed file right now so... yeah...
data("stomach_data_fit")
stf<-stomach_data_fit
sp  <- species_params(spawning_mort)
idx <- match(sp$Scientific_name, stf$species)  # align order to species_params

lambda <- 2
sp$pred_kernel_type <- c("power_law","lognormal","power_law","power_law","power_law","power_law","power_law","power_law","power_law","power_law","power_law","lognormal","power_law","power_law")
sp$kernel_exp <- stf$alpha[idx] + 4/3 - lambda
sp$kernel_l_l <- stf$ll[idx]
sp$kernel_u_l <- stf$ul[idx]
sp$kernel_l_r <- stf$lr[idx]
sp$kernel_u_r <- stf$ur[idx]

species_params(spawning_mort) <- sp
```

Continuous re-tuning

```{r}
spawning_mort<-readRDS("/Users/jessicawestworth/Desktop/spawning_mort.rds")
blue_whiting<-tuneEcopath(
    spawning_mort, 
    catch = all_fishing,
    age_at_length = cs_age_size,
    diet = dm)

#saveRDS(blue_whiting,"/Users/jessicawestworth/Desktop/spawning_mort.rds")
```
