---
title: "Celtic_Sea_Status_Quo_Model"
author: "Jessica Westworth"
format: pdf
editor: visual
---

# Introduction

Creating a model representing the average state of the Celtic Sea from 2012 to 2024. Composed of 15 species: herring (*Clupea harengus*), cod (*Gadus morhua*), haddock (*Melanogrammus aeglefinus*), whiting (*Merlangius merlangus*), blue whiting (*Micromesistius poutassou*), hake (*Merluccius merluccius*), monkfish (Lophius piscatorius), horse mackerel (*Trachurus trachurus*), mackerel (*Scomber scombrus*), plaice (*Pleuronectes platessa*), megrim (*Lepidorhombus whiffiagonis*), sole (*Solea solea*), boarfish (*Capros aper*), dragonet (Callionymus spp.), and red gurnard (Chelidonicthys cuculus).

## Loading in Repository and Libraries

```{r}
#| warning: false
#| message: false
#| results: 'hide'

#library(remotes)
#install_github("sizespectrum/mizer", ref="diffusion") 
#install_github("gustavdelius/mizerEcopath", ref="spawning_season") 
library(mizer)
library(mizerEcopath)
library(dplyr)
library(here)
```

## Load in Data

Data includes: life_history parameters (l_mat and age_mat calculated from DATRAS data; w_max and LWR from Fishbase); landing distributions form STECF; survey size distribution and age at length distribution data from DATRAS; Stomach contents data; Biomass estimates, yield estimates, interaction matrix, and diet matrix from Lauria (2016).

```{r}
#| warning: false
#| message: false
#| results: 'hide'

#Life history data from fishbase
data(life_history_fishbase)

#maturity values calculated from survey data
data(cs_calculated_histories)

    #combine life_history_fishbase and calculated l_mat
life_histories<-life_history_fishbase%>%
    left_join(cs_calculated_histories, by="Scientific_name")%>%
    mutate(
        l_mat=case_when(
        l_mat.y %in% c(NA_character_) ~ l_mat.x,
        TRUE ~ l_mat.y),
        age_mat=case_when(
        age_mat.x %in% c(NA_character_) ~ age_mat.y,    
        TRUE ~ age_mat.x))%>%
    arrange(species)%>%
    select(-l_mat.x, -l_mat.y, -age_mat.x, -age_mat.y)

#Load in ecopath data
ecopath_estimates <- read.csv(here("inst","extdata","Ecopath-Basic estimates.csv"))
    #Biomass
observed_biomass_ecopath<-ecopath_estimates[c('Group.name','Biomass..t.km.2.')]%>%
    mutate(species= Group.name,
           biomass_observed = Biomass..t.km.2.)%>%
    select(-Group.name,-Biomass..t.km.2.)%>%
    filter(species %in% c("Blue whiting","Boarfish","Cod","Dragonet",
                          "Haddock","Hake","Herring","Horse mackerel",
                          "Mackerel","Megrim","Monkfish","Plaice",
                          "Red gurnard","Sole","Whiting"))
observed_biomass_ecopath<-na.omit(observed_biomass_ecopath)

#Biomass from literature for Red gurnard, Boarfish, and dragonets
Boarfish<-data.frame(species="Boarfish", biomass_observed=1.4126)
Dragonet<-data.frame(species="Dragonet", biomass_observed=0.0289)
Red_gurnard<-data.frame(species="Red gurnard", biomass_observed=0.0058)
#combine the ecopath species and additional species
observed_biomasses<-rbind(observed_biomass_ecopath,Boarfish,Dragonet,Red_gurnard)
observed_biomasses<-arrange(observed_biomasses,species)

#Diet matrix
data(diets)

#Deaths from fishing
data(fishing_deaths)

#Catch Distribution
data(catch_distribution)

#DATRAS size distribution
data(survey_length_distribution)

#DATRAS age at length distribution
data(cs_age_size)

#determining l_max comparing landings max length versus fishbase max length recorded value
observed_max <- catch_distribution %>%
     group_by(species) %>%
     summarise(
         obs_len = if (all(is.na(length + dl))) NA_real_ else max(length + dl, na.rm = TRUE)
     )
life_histories <- life_histories %>%
    left_join(observed_max, by = "species")%>%
    mutate(
        l_max = pmax(obs_len, l_max))%>%
    select(-obs_len)

#combined survey fishing and STECF fishing
all_fishing<-rbind(catch_distribution,survey_length_distribution)
```

## Species Parameters

Assign species parameters from life history data

```{r}
#calculate w_mat and w_max and set n,p,d, and alpha parameter
sp_params<-life_histories%>%
    mutate(w_mat=a*l_mat^b,
           w_max=a*l_max^b,
           n=0.7,
           p=0.7,
           d=-0.3,
           alpha=0.7,
           biomass_cutoff=w_mat)%>%
    left_join(observed_biomasses, by="species")
```

## Build Non-Interacting Model

```{r}
model_non_int <- newAllometricParams(sp_params, no_w = 200)

## Add gear params
model_non_int <- addCatch(model_non_int, landings= catch_distribution,
                              fishing_dead_biomass=fishing_deaths,
                              survey= survey_length_distribution,step=2)

# Steady model
model_non_int <- model_non_int %>%
    steadySingleSpecies() %>%
    matchBiomasses() %>%
    steadySingleSpecies() %>%
    matchBiomasses()%>%
    steadySingleSpecies() %>%
    matchBiomasses()%>%
    steadySingleSpecies() %>%
    matchBiomasses()%>%
    steadySingleSpecies() %>%
    matchBiomasses()

model_non_int_1<-tuneEcopath(
    model_non_int, 
    catch = all_fishing,
    age_at_length = cs_age_size)
```
