---
title: "Understanding and Extending the Tuning Gadget"
author: "Gustav Delius"
format:
  html:
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{Understanding and Extending the Tuning Gadget}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
# The code chunks in this article define Shiny UI/server functions and launch
# interactive gadgets, so we do not evaluate them when the article is built.
knitr::opts_chunk$set(eval = FALSE)
```

## Introduction

mizerEcopath ships two interactive tools for tuning a model by hand:

- `tuneParams()` has a **"steady"** button. Each time you press it the model is
  projected to its multi-species steady state with `mizer::steady()`. This tool
  is meant for hand-tuning fully interacting, multi-species models.
- `tuneEcopath()` has a **"match"** button. Each time you press it a set of
  optimisers is run to match the model to Ecopath data. This tool is meant for
  non-interacting models with a fixed growth rate and a power-law mortality.

Although the two serve different workflows, they are *not* two separate apps.
Both are thin wrappers around a single shared engine, `tuningGadget()`. The
sidebar, the tabbed main panel, the species selector, the undo/redo log, the
flow of the reactive parameter object, and the action button are all defined
once, in `tuningGadget()`. The two functions differ only in the configuration
they hand to that engine:

| | `tuneParams()` | `tuneEcopath()` |
|---|---|---|
| action button | "steady" (shortcut `s`) | "match" (shortcut `m`) |
| what the button does | `mizer::steady()` | `matchGrowth()`, `matchCatch()`, ... |
| controls shown | `predation`, `fishing`, ... | `ecopathGrowth`, `ecopathFishing`, ... |
| tabs shown | `Spectra`, `Growth`, `Death`, ... | `ecopathSpectra`, `ecopathGrowth`, ... |

This article explains how that engine works so that you can adapt it to your own
workflow: pick and choose which controls and tabs to show, write your own
controls and tabs, and — for the more adventurous — call `tuningGadget()`
directly with your own action button.

## Anatomy of the gadget

The gadget has two parts:

- A **sidebar** containing the action button, the undo/redo buttons, the
  `Match:` checkboxes, the species selector, and a stack of *parameter
  controls*. Each control is a small block of input widgets (usually sliders)
  that change the parameters of the currently selected species.
- A **main panel** containing a set of *tabs*. Each tab shows one or more
  diagnostic plots computed from the current model.

Controls and tabs are the two kinds of plug-in. You choose which ones appear
through the `controls` and `tabs` arguments, and you can add your own by
following a naming convention described below.

## The simplest customisation: choosing controls and tabs

Both `controls` and `tabs` are plain character vectors. To show fewer panels,
pass a shorter vector; to reorder them, reorder the vector. For example, a
stripped-down growth-tuning session:

```{r}
tuneParams(params,
           controls = c("growth"),
           tabs = c("Growth", "Spectra"))
```

This is exactly how the convenience wrapper `tuneGrowth()` is defined — it is
just a preset of `tuneParams()` with a short `controls`/`tabs` list.

Each name in these vectors is resolved to a pair of functions by a naming
convention. That convention is the key to writing your own panels.

## The naming convention

For a **control** named `"foo"` the engine expects two functions:

- `fooControlUI(p, input)` — returns the input widgets for the sidebar.
- `fooControl(input, output, session, params, params_old, flags, trigger_update, ...)`
  — the server logic that reacts to those widgets.

The engine renders the section heading for you, so `fooControlUI()` should
**not** produce its own heading — it returns only its input widgets. The heading
text is taken from an optional string `fooControlTitle`; if you do not define one
it defaults to the control name. You may also define a string
`fooControlDescription`, which is shown as a tooltip both on the heading and on
the sidebar link that jumps to the control. The heading carries an anchor whose
id is the control name, so those sidebar links (and any `href="#foo"` links you
put in a tab's help text) scroll to it.

Control names should start with a **lower-case** letter (the UI function name is
used verbatim, while the server function name has its first letter lower-cased
internally).

For a **tab** named `"Foo"` (or `"foo"`) the engine expects:

- `fooTabUI(params, ...)` — returns the tab layout.
- `fooTab(input, output, session, params, params_old, logs, trigger_update, ...)`
  — the server logic that produces the tab's outputs.

Optionally define a string `fooTabTitle` to label the tab; otherwise the label
is the name you passed. The engine lower-cases the first letter of a tab entry
before looking up its functions, so your functions must start with a lower-case
letter (`fooTab`, not `FooTab`), but you may still pass a capitalised entry like
`"Foo"` to get a nicer default tab label.

The functions are found by name on the search path, so **defining them in your
global environment is enough** — you do not need to add them to a package.

## Writing a new tab

The package ships a template you can copy, `exampleTab.R`. Here is its essence:

```{r}
exampleTabUI <- function(...) {
    tagList(
        plotlyOutput("plot_example"),
        radioButtons("example_prop", "Show",
                     choices = c("Proportion", "Rate"),
                     selected = "Proportion",
                     inline = TRUE)
    )
}

exampleTab <- function(input, output, session, params, logs, ...) {
    output$plot_example <- renderPlotly({
        plotlyDeath(params(),
                    species = input$sp,
                    proportion = (input$example_prop == "Proportion"))
    })
}

# Optional: the label shown on the tab rider
exampleTabTitle <- "E.g."
```

A few things to note about the signatures:

- The UI function receives **only `params`** (a Shiny `reactiveVal`). If you need
  the current model to lay out the UI, read it with `isolate(params())`. If you
  do not, you can just write `function(...)` as above.
- The server function receives the reactive `params` plus the other reactive
  machinery. Inside a `render*()` block, call `params()` (reactively) to get the
  current model; it re-renders automatically whenever the model changes.
- `input$sp` always holds the name of the currently selected species, so most
  tabs plot the selected species.

To use it, list its name alongside the built-in tabs:

```{r}
tuneParams(celtic_params, tabs = c("Spectra", "Growth", "Example"))
```

### A worked example

Suppose we want a tab that plots the growth curve of the selected species next
to the community size spectrum. We define the UI and server functions and pass
the tab's name to `tuneParams()`:

```{r}
myGrowthTabUI <- function(...) {
    tagList(
        plotOutput("my_growth_curve"),
        plotlyOutput("my_spectrum")
    )
}

myGrowthTab <- function(input, output, session, params, ...) {
    output$my_growth_curve <- renderPlot({
        plotGrowthCurves(params(), species = input$sp)
    })
    output$my_spectrum <- renderPlotly({
        plotSpectra(params(), power = 2, highlight = input$sp)
    })
}

myGrowthTabTitle <- "My growth"

tuneParams(celtic_params, tabs = c("Spectra", "myGrowth"))
```

Because these functions live in the global environment, the engine finds them by
name — no package installation required.

## Writing a new control

A control changes model parameters. Its UI function builds the widgets from the
*current* value of the parameters for the selected species, and its server
function reacts to those widgets and pushes the changed model back into the app.

The built-in `growthControl` is a good model to imitate:

```{r}
growthControlUI <- function(p, input) {
    sp <- p@species_params[input$sp, ]
    tagList(
        sliderInput("gamma", "Search volume coefficient 'gamma'",
                    value = sp$gamma,
                    min = signif(sp$gamma / 2, 3),
                    max = signif(sp$gamma * 1.5, 3),
                    step = sp$gamma / 50, ticks = FALSE)
    )
}

growthControl <- function(input, output, session, params, params_old, flags, ...) {
    observeEvent(input$gamma, {
        p <- params()
        sp <- input$sp
        # Skip the first fire after a species switch (see "The flags pattern")
        if (!identical(sp, flags$sp_old_pred)) {
            flags$sp_old_pred <- sp
            return()
        }
        p@species_params[sp, "gamma"] <- input$gamma
        p <- setSearchVolume(p)
        tuneParams_update_species(sp, p, params, params_old)
    },
    ignoreInit = TRUE, ignoreNULL = TRUE)
}
```

Note the differences from a tab:

- The **UI function receives `p`** — a plain, already-isolated `MizerParams`
  object — and `input`. You read the selected species' parameters directly from
  `p@species_params[input$sp, ]` to set the slider's initial value and range.
- The **UI function returns only its widgets** — no heading. The engine adds the
  section heading (from `growthControlTitle`, or the control name) above them, so
  you never write a `tags$h3()` of your own. A control may still include *inner*
  sub-headings for grouping (as `otherControl` does with its "Mort" heading); it
  is only the top-level section heading that the engine owns.
- The **server function receives the reactive `params` and `params_old`**, plus
  a `flags` environment. After changing a parameter you commit the change with
  one of the helper functions described below.

### The `flags` pattern

There is a subtlety. When you switch to a different species, the sidebar sliders
are rebuilt to show that species' values. Rebuilding a slider fires its
`observeEvent`, which — if you did nothing to prevent it — would immediately
write the newly displayed value back into the model, potentially corrupting it.

To avoid this, controls stash the last species they saw in the shared `flags`
environment and bail out on the first fire after a switch:

```{r}
if (!identical(sp, flags$sp_old_pred)) {
    flags$sp_old_pred <- sp
    return()
}
```

Use a **unique flag name per control** (here `sp_old_pred`) so controls do not
interfere with one another.

### A worked example

Here is a self-contained control that adds a slider for the maximum intake rate
coefficient `h`:

```{r}
intakeControlUI <- function(p, input) {
    sp <- p@species_params[input$sp, ]
    tagList(
        sliderInput("h", "Max intake coefficient 'h'",
                    value = sp$h,
                    min = signif(sp$h / 2, 3),
                    max = signif(sp$h * 1.5, 3),
                    step = sp$h / 50, ticks = FALSE)
    )
}

intakeControl <- function(input, output, session, params, params_old, flags, ...) {
    observeEvent(input$h, {
        p <- params()
        sp <- input$sp
        if (!identical(sp, flags$sp_old_intake)) {
            flags$sp_old_intake <- sp
            return()
        }
        p@species_params[sp, "h"] <- input$h
        p <- setMaxIntakeRate(p)
        tuneParams_update_species(sp, p, params, params_old)
    },
    ignoreInit = TRUE, ignoreNULL = TRUE)
}

# Heading shown above the control (without this it would default to "intake")
intakeControlTitle <- "Intake"
# Optional tooltip shown on the heading and on the sidebar link to the control
intakeControlDescription <- "Adjust the maximum intake rate coefficient 'h'."

tuneParams(celtic_params, controls = c("growth", "intake"))
```

## The reactive state and the update helpers

Understanding two reactive values makes the rest of the framework fall into
place.

- **`params`** — a `reactiveVal` holding the model as it is currently displayed.
  Every plot reads from it; every control writes to it.
- **`params_old`** — a `reactiveVal` holding the abundances of the last
  *committed* steady state. It serves as a fixed reference community, as
  explained next. It is updated only when you press the action button or move
  through the undo log, never by an ordinary slider change.

### Single-species updates and the frozen environment

To make sense of `params_old` you first have to understand what happens when you
change a parameter.

When you drag a slider you want to see the effect at once. Computing the true
multi-species steady state — what the action button does — is far too slow to
run on every slider movement. So while you tune, the gadget does something
cheaper and deliberately more local. When you change a parameter of the selected
species, it recomputes **only that species' steady-state size spectrum**, and it
does so from *frozen rates*. Before anything is recomputed, the abundances of
every species and of the resource — including the selected species itself — are
set back to the previous state. The growth and mortality rates for the selected
species are then evaluated from that state and held fixed while its new spectrum
is solved for.

This matters because the rates a fish experiences are not a property of that fish
alone. Its growth rate depends on how much prey is available and its death rate
on how many predators there are, and those depend on the abundances of the whole
community — the other species, the resource, and the selected species itself
(through cannibalism or competition for shared prey, for example). Evaluating
every rate at the previous state freezes all of this at once. In particular it
freezes the selected species' own contribution: within a single update its
recomputed spectrum does not feed back into the rates that produced it. The
update simply answers one question — holding every rate at what it was in the
previous state, what spectrum would this species relax to?

The consequence is that **there is no feedback**. If you make the selected
species grow faster, you immediately see its own spectrum shift, but its prey are
not depleted in response and its predators do not become more numerous. In
reality all of those would change, and would in turn feed back on the focal
species. Those multi-species feedbacks are worked out only when you press the
action button, which projects the whole community forward together until
everything settles into a mutually consistent steady state.

In a *non-interacting* model — the kind tuned by `tuneEcopath()` — species do
not prey on one another, so each species always lives in a fixed background.
There the single-species update is not an approximation but the whole story, and
the action button's job is to match the model to data rather than to resolve any
community feedback.

This is exactly what `params_old` is for. The frozen environment has to be a
*fixed reference*. If instead each recomputation started from the currently
displayed model, the environment would drift as you worked: dragging a slider
one way and then back would not return you to where you started, because each
update would build on the perturbed result of the last one, and small changes
would silently accumulate. To prevent this, every single-species update first
resets all abundances to those stored in `params_old` — the abundances of the
last committed steady state — and only then re-solves the focal species. So
`params_old` holds the reference community against which all your in-between
tweaks are measured, and it changes only when you deliberately commit a new
steady state (via the action button) or step through the undo/redo log.

The helper `tuneParams_update_species()` performs this whole dance for you:

```{r}
# inside tuneParams_update_species(sp, p, params, params_old):
p@initial_n <- params_old()@initial_n   # reset ALL abundances (incl. the focal
                                        # species) to the previous state
p <- steadySingleSpecies(p)             # evaluate rates from that state, then
                                        # re-solve only the focal species
tuneParams_update_params(p, params)     # display the result
```

so in your own controls you normally just call it and do not manipulate the
abundances yourself.

### The helper functions

You do not manipulate the reactive values directly. Instead you call one of the
exported helper functions, choosing the one that matches your intent:

| Helper | Use it when |
|---|---|
| `tuneParams_update_params(p, params)` | You changed the model directly and want to display it (marks the state as changed, enables Undo). No steady-state calculation. Typical in **tabs**. |
| `tuneParams_update_species(sp, p, params, params_old)` | You changed a growth/encounter parameter and want that species re-solved to its single-species steady state before display. Typical in **controls**. |
| `tuneParams_update_abundance(p, sp, params, params_old)` | You edited a species' abundance directly and want the change to persist (it also writes into `params_old`). |
| `tuneParams_add_to_logs(logs, p, params)` | You want to commit a state *and* push an entry onto the undo log. Typical in **action handlers**. |
| `error_fun(e)` | The standard "invalid parameters" modal, for use as the `error` argument of `tryCatch()`. |

Two more pieces of shared state are occasionally useful:

- **`trigger_update`** — a `reactiveVal`. Calling `trigger_update(runif(1))`
  forces the sidebar sliders to rebuild. Do this after an action that changes
  parameters behind the sliders' backs (for example, an optimiser), so the
  displayed slider values catch up.
- **`logs`** — an environment holding the on-disk undo/redo log. You rarely
  touch it except by passing it to `tuneParams_add_to_logs()`.

## The action button

The single large button in the sidebar (the "steady" button in `tuneParams()`,
the "match" button in `tuneEcopath()`) is fully configurable. Four arguments to
`tuningGadget()` define it:

- `action_label` — the button's HTML label, e.g. `HTML("<u>s</u>teady")`.
- `action_tooltip` — the hover text.
- `action_key` — the JavaScript key code of its keyboard shortcut (`83` for `s`,
  `77` for `m`).
- `action_handler` — the function that runs when it is pressed.

The handler has the signature

```{r}
function(p, input, session, params, params_old, logs, flags, trigger_update, ...)
```

where `p` is the current model. A handler typically transforms `p`, then commits
the result. The steady handler is essentially

```{r}
tuneParams_steady_handler <- function(p, input, session, params, params_old,
                                      logs, flags, trigger_update, ...) {
    p <- mizer::steady(p)
    params_old(p)
    tuneParams_add_to_logs(logs, p, params)
}
```

while the Ecopath match handler runs a sequence of `match*()` optimisers instead
of `steady()`.

### The `Match:` checkboxes

The checkboxes above the species selector are wired to the action button through
`input$match`. You define which options appear and which are ticked by default:

```{r}
tuningGadget(
    ...,
    match_choices  = c("growth", "biomass", "yield"),
    match_selected = c("none")
)
```

Your `action_handler` reads `input$match` to decide what to do. For example the
steady handler optionally calls `matchGrowth()`, `matchBiomasses()` or
`matchYield()` depending on which boxes are ticked before running to steady
state.

## Startup and shutdown hooks

Three optional hooks let you inject package-specific behaviour without touching
the engine:

- `prepare_params_hook(p) -> p` runs once at launch, after the shared
  `prepare_params()`. Use it to set defaults or derived parameters your controls
  and tabs assume. `tuneEcopath()` uses it to set encounter, gonad and diffusion
  defaults and to run an initial `steadySingleSpecies()`.
- `finalise_params_hook(p) -> p` runs when the user presses "Return" or
  downloads the model. Use it to undo any temporary setup. `tuneParams()` uses it
  to remove a temporary "no gear" gear it added in its prepare hook.
- `update_species_hook(sp, p) -> p` runs inside
  `tuneParams_update_species()`, right after `steadySingleSpecies()`, every time
  a single species is recomputed. `tuneEcopath()` uses it to call
  `matchBiomasses()` so that biomass is preserved as parameters are tuned.

## Building a fully custom gadget

Because "steady" and "match" are just two instances of the same action
mechanism, you can build a third variant by calling `tuningGadget()` directly.
This is how `tuneParams()` itself is written:

```{r}
myTuner <- function(params, ...) {
    tuningGadget(
        params = params,
        controls = c("growth", "intake"),
        tabs = c("Spectra", "myGrowth", "Death"),
        match_choices = c("growth", "biomass"),
        match_selected = c("none"),
        action_label = HTML("<u>s</u>teady"),
        action_tooltip = "Find steady state. Keyboard shortcut: s",
        action_key = 83,
        action_handler = tuneParams_steady_handler,
        ...
    )
}
```

You can supply extra arguments that your tabs need through `tab_extra_args` (a
named list passed to every tab server function) — this is how `tuneEcopath()`
hands the observed `catch` and `diet` data to its tabs.

## Summary

- `tuningGadget()` is the shared engine; `tuneParams()` and `tuneEcopath()` are
  presets that differ only in configuration.
- Show or hide panels by editing the `controls` and `tabs` vectors.
- Add a **tab** by defining `fooTabUI(params, ...)` and
  `fooTab(input, output, session, params, ...)` (see `exampleTab.R`).
- Add a **control** by defining `fooControlUI(p, input)` and
  `fooControl(input, output, session, params, params_old, flags, ...)`,
  remembering the `flags` guard and committing changes with
  `tuneParams_update_species()`. Return only the widgets: the engine renders the
  heading from `fooControlTitle` (or the control name), with an optional
  `fooControlDescription` tooltip.
- Commit changes with the right helper: `tuneParams_update_params()`,
  `tuneParams_update_species()`, `tuneParams_update_abundance()` or
  `tuneParams_add_to_logs()`.
- Replace the whole action button by passing your own `action_handler` (and
  `action_label`/`action_key`) to `tuningGadget()`.

## See also

`tuningGadget()`, `tuneParams()`, `tuneEcopath()`, `tuneGrowth()`, and the
reference index sections "gadget tabs" and "gadget controls" for the full list
of built-in panels you can imitate.
