---
title: "Calculate Biomass Discard Rate"
format: html
editor: visual
---

## Importing and curating the data

```{r}
# Load required libraries
library(tidyverse)
library(here)
library(mizer)
library(here)
library(mizerExperimental)

#Need to check if we can upload these to the reprository, if so, we can make this generalised and take from the reprository rather than my machine.
FDI_Discards_Length<-read.csv("/Users/jessicawestworth/Desktop/FDI Discards Length.csv")
species_codes <- read.csv("/Users/jessicawestworth/Desktop/ASFIS_sp_2022.csv")
```

## Clean Data

```{r}
library(dplyr)
library(stringr)
library(purrr)

#Remove rows with value NA from the no length column
FDI_Discards_Length_clean <- FDI_Discards_Length[!is.na(FDI_Discards_Length$tot_discards_tonnes), ]
FDI_Discards_Length_clean <- FDI_Discards_Length_clean[!is.na(FDI_Discards_Length$total_live_weight_landed), ]

FDI_Discards_Length_clean <- FDI_Discards_Length_clean %>%
  mutate(
    domain_discards = if_else(
      country == "Spain" & !startsWith(domain_discards, "ESP_"),
      paste0("ESP_", domain_discards),
      domain_discards
    )
  )

#Remove NIR
FDI_Discards_Length_clean <- FDI_Discards_Length_clean %>%
  filter(!str_starts(domain_discards, "NIR"))

# For discards
FDI_Discards_Length_clean <- FDI_Discards_Length_clean %>%
  mutate(domain_discards = domain_discards %>%
             str_replace_all("_EU_", "_")%>%
             str_replace_all("EU_", "_")%>%
             str_replace_all("_COAST_", "_")%>%
             str_replace_all("COAST_", "_")%>%
             str_replace_all("_UK_", "_"))
```

## Configurable parameters - subareas and gears

The following script is used for the Celtic Sea, however, by defining different subareas one could extract data for a variety of other areas.

```{r}
sub_area_7<- c('27.7.B', '27.7.C', '27.7.E', '27.7.F', '27.7.G', '27.7.H', '27.7.J', '27.7.K')
ns_sub_area<-c('27.4.A','27.4.B','27.4.C') #potentially add 27.4 to this list?
```

## Data preparation

```{r}
# This function extracts gear types *only* from the 4th underscore-separated domain part, ensure that the domains follow the pattern till gear type detailed above before running.
extract_gear_types_combined <- function(domain_parts) {
  if (length(domain_parts) < 4) return(NA_character_)
  gear_block <- domain_parts[4]
  gear_codes <- str_split(gear_block, "-", simplify = FALSE)[[1]]
  paste(gear_codes, collapse = "-")
}

extract_target_assemblage_combined <- function(domain_parts) {
  if (length(domain_parts) < 5) return(NA_character_)
  gear_block <- domain_parts[5]
  gear_codes <- str_split(gear_block, "-", simplify = FALSE)[[1]]
  paste(gear_codes, collapse = "-")
}

length_discards <- FDI_Discards_Length_clean %>%
  rowwise() %>%
  mutate(
    domain_parts = list(str_split(domain_discards, "_")[[1]]),
    region_code = domain_parts[3],  # e.g. 27.4.a-27.4.b
    gear_type = extract_gear_types_combined(domain_parts),
    target_assemblage= extract_target_assemblage_combined(domain_parts),
    
    # Normalize lengths
    length = ifelse(length_unit == "mm", length / 10, length),
    length_unit = "cm",
    source = "discards"
  ) %>%
  ungroup() %>%
  select(-domain_parts) %>%
  left_join(species_codes %>% 
              select(X3A_CODE, Scientific_name, English_name),
            by = c("species" = "X3A_CODE")) %>%
  rename(domain = domain_discards) %>%
  select(
    country, year, domain, species, Scientific_name, English_name,
    length_unit, region_code, gear_type, target_assemblage,
    no_length_measurements, length, total_live_weight_landed, tot_discards_tonnes,
    source
  )
```

## Subarea filtering

Use the previously defined regions above in the configurable parameters section and extract them from the region_code column in the landings_length data frame. This will create a new column called sub_region and give them the value cs_sub area if the values contain any combinations of the defined areas, but reject any rows that contain other sub areas in addition to the defined regions.

Example of treatment:

| region_code   | sub_region  |
|---------------|-------------|
| 27.7.b        | cs_sub_area |
| 27.7.b-27.7.h | cs_sub_area |
| 27.7.g-27.7.a | NA          |
| 27.4.a        | ns_sub_area |
| 27.4.b-27.4.c | ns_sub_area |
| 27.4.a-27.7.h | NA          |

```{r}
length_landings <- length_landings %>%
  mutate(
    # Standardize region_code format (uppercase)
    region_code_upper = toupper(region_code),

    # Split region_code into individual codes (handling multi-code entries)
    region_parts = str_split(region_code_upper, "-"),

    # Assign sub_region if ALL codes in a row match one region group
    sub_region = case_when(
      sapply(region_parts, function(rcs) all(rcs %in% sub_area_7)) ~ "cs_sub_area",
      sapply(region_parts, function(rcs) all(rcs %in% ns_sub_area)) ~ "ns_sub_area",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-region_code_upper, -region_parts)  # Clean up helper columns

length_discards <- length_discards %>%
  mutate(
    # Standardize region_code format (uppercase)
    region_code_upper = toupper(region_code),

    # Split region_code into individual codes (handling multi-code entries)
    region_parts = str_split(region_code_upper, "-"),

    # Assign sub_region if ALL codes in a row match one region group
    sub_region = case_when(
      sapply(region_parts, function(rcs) all(rcs %in% sub_area_7)) ~ "cs_sub_area",
      sapply(region_parts, function(rcs) all(rcs %in% ns_sub_area)) ~ "ns_sub_area",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-region_code_upper, -region_parts)  # Clean up helper columns
```

# Celtic Sea

### Subset data for Celtic Sea

```{r}
cs_sub_region_discards<-subset(length_discards, sub_region=="cs_sub_area")
```

### Calculate weight by species, size, and gear type

```{r}
# Define a function to calculate catch by species, size class, and gear type
weight_by_size_class_and_gear <- function(data) {
    data %>%
        group_by(species, length, gear_type, target_assemblage, country, source) %>%
        summarise(landed_tonnes= sum(total_live_weight_landed),
                  discards_tonnes = sum(tot_discards_tonnes)) %>%
        ungroup()
}

cs_discards_data <- weight_by_size_class_and_gear(cs_sub_region_discards)

species_reference_discards <- length_discards %>%
    distinct(species, Scientific_name, English_name)
```

### Add species names and map gear categories

```{r}
cs_discards_data <- cs_discards_data %>%
  left_join(species_reference_discards, by = "species") %>%
  rename(
    gear = gear_type,  # Adjust as needed if `gear_type` is used elsewhere
    landed_tonnes = landed_tonnes,
    discards_tonnes = discards_tonnes,
    target_assemblage =target_assemblage,
    country=country
  ) %>%
  drop_na()  # Remove rows with NA in any column
```

## Generating a size-distributed catch data frame

### Species filter

cs_discards

```{r}
subset_species <- function(data, species_list, match_by = "species") {
  # Validate match_by parameter
  if (!match_by %in% c("species", "Scientific_name", "English_name")) {
    stop("match_by must be one of 'species', 'Scientific_name', or 'English_name'")
  }
  
  # Filter the data to include only the specified species
  subset_data <- data %>% filter(!!sym(match_by) %in% species_list)
  
  return(subset_data)
}

species_list <- c("HER","COD","HAD","WHG","WHB","HKE","MON","HOM","MAC","PLE","MEG","SOL","GUR", "BOC", "LYY")

cs_discards <- subset_species(
  data = cs_discards_data,
  species_list = species_list,
  match_by = "species"
 )

# Rename the species column to species_code
colnames(cs_discards)[colnames(cs_discards) == "species"] <- "species_code"

# Create a named vector for the new species names
new_species_names <- c("Atlantic herring"="Herring",
"Atlantic cod"="Cod", 
"Haddock"="Haddock",
"Whiting"="Whiting",
"Blue whiting(=Poutassou)"="Blue whiting",
"European hake"="Hake",
"Angler(=Monk)"="Monkfish",
"Atlantic horse mackerel"="Horse mackerel",
"Atlantic mackerel"="Mackerel",
"European plaice"="Plaice",
"Megrim"="Megrim",
"Common sole"="Sole",
"Boarfish"="Boarfish",
"Dragonet"="Dragonet",
"Red gurnard"="Red gurnard"
)

# Add the new species column by matching English_name
cs_discards$species <- new_species_names[cs_discards$English_name]

# Add the dl column
cs_discards$dl <- 1

#convert catch to per year per area (m^2): total area surveyed is area of celtic sea 6.174997e+11 (m^2) and number of years is 2013 to 2020 so 8 years
#cs_landings$catch<-cs_landings$catch/6.174997e+11/8
#saveRDS(cs_catch, here("inst","extdata","cs_catch.rds"))
#cs_discards$catch<-cs_discards$catch/6.174997e+11/8
```

Combining gears to ecopath gears

```{r}
#ecopath gear groups?
cs_discards <- cs_discards %>%
  mutate(
    gear_category = case_when(
      gear %in% c("OTT","OTB","OTB-OTT-PTB","TBB") ~ "Trawl",
      gear %in% c("GND-GNS-GNC","GNS","GTR","GTR-GTN") ~ "Gillnet",
      gear %in% c("LHM-LHP","OTM","OTM-PTM","LLS","PS","PS-LA","SSC") ~ "Pelagic",
      TRUE ~ "Misc") # default for everything else
  )


cs_discards <- cs_discards %>%
  group_by(species, country, Scientific_name, length,gear_category, target_assemblage, source) %>%
  summarise(
    landed_tonnes = sum(landed_tonnes, na.rm =TRUE),
    discards_tonnes = sum(discards_tonnes, na.rm = TRUE),
    .groups = "drop")
```

Calculate discard percentage for each species, country, length, gear, and target assemblage

```{r}
cs_discards_dictionary<-cs_discards%>%
    mutate(discard_biomass_rate=case_when(
        landed_tonnes == 0 ~ 1,
        TRUE ~ discards_tonnes/landed_tonnes)
    )
    
```

```{r}
FDI_EU_2013<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2013.csv")
FDI_EU_2014<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2014.csv")
FDI_EU_2015<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2015.csv")
FDI_EU_2016<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2016.csv")
FDI_EU_2017<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2017.csv")
FDI_EU_2018<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2018.csv")
FDI_EU_2019<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2019.csv")
FDI_EU_2020<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2020.csv")
FDI_EU_2021<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2021.csv")
FDI_EU_2022<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2022.csv")
FDI_EU_2023<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2023.csv")

FDI_EU<-rbind(FDI_EU_2013,FDI_EU_2014,FDI_EU_2015,FDI_EU_2016,FDI_EU_2017,FDI_EU_2018,FDI_EU_2019,FDI_EU_2020,FDI_EU_2021,FDI_EU_2022,FDI_EU_2023)

unique(FDI_EU$sub_region)

FDI_EU <- FDI_EU %>%
  mutate(
    sub_region = case_when(
      sub_region %in% c("27.7.B") ~ "cs",
      sub_region %in% c("27.7.D") ~ "cs",
      sub_region %in% c("27.7.E") ~ "cs",
      sub_region %in% c("27.7.F") ~ "cs",
      sub_region %in% c("27.7.G") ~ "cs",
      sub_region %in% c("27.7.H") ~ "cs",
      sub_region %in% c("27.7.J") ~ "cs",
      TRUE ~ NA_character_),
    total_live_weight_landed = case_when(
        total_live_weight_landed %in% c("C") ~ NA_character_,
        TRUE ~ total_live_weight_landed),
    tot_discards_tonnes = case_when(
        tot_discards_tonnes %in% c("C") ~ NA_character_,
        TRUE ~ tot_discards_tonnes))%>%
    filter(!is.na(total_live_weight_landed),
           !is.na(tot_discards_tonnes))
    

FDI_EU<-subset(FDI_EU, sub_region=="cs")
FDI_EU$tot_discards_tonnes<-as.numeric(FDI_EU$tot_discards_tonnes)
FDI_EU<-FDI_EU[c("gear_type","species","country","target_assemblage","total_live_weight_landed","tot_discards_tonnes")]
FDI_EU$tot_discards_tonnes<-replace_na(FDI_EU$tot_discards_tonnes, 0)
FDI_EU$total_live_weight_landed<-as.numeric(FDI_EU$total_live_weight_landed)
FDI_EU<-na.omit(FDI_EU)

FDI_EU<-FDI_EU %>%
    group_by(gear_type, species,country,target_assemblage)%>%
    summarise(total_landings = sum(total_live_weight_landed),
              tot_discards_tonnes = sum(tot_discards_tonnes))

species_list <- c("HER","COD","HAD","WHG","WHB","HKE","MON","HOM","MAC","PLE","MEG","SOL","GUR", "BOC", "LYY")

FDI_EU <- subset_species(
  data = FDI_EU,
  species_list = species_list,
  match_by = "species"
 )

FDI_EU <- FDI_EU%>%
    mutate(
        gear_category = case_when(
             gear_type %in% c("OTT","OTB","OTB-OTT-PTB","PTB","TBB") ~ "Trawl",
             gear_type %in% c("GND-GNS-GNC","GNS","GTR","GTR-GTN",
                               "GND","GTN","GNC") ~ "Gillnet",
             gear_type %in% c("LNB","LTL","LHM-LHP","LHM","LHP","LLD","OTM","OTM-PTM",
                              "PTM","SB","SDN","SV","SPR","LLS","PS","PS-LA",
                              "SSC","GEF") ~ "Pelagic",
             TRUE ~ "Misc"))%>% # default for everything else
    group_by(gear_category, species, country, target_assemblage)%>%
    summarise(total_landings = sum(total_landings),
              tot_discards_tonnes =sum(tot_discards_tonnes))%>%
    mutate(discard_rate = tot_discards_tonnes/total_landings)

new_species_names <- c("HER"="Herring",
"COD"="Cod", 
"HAD"="Haddock",
"WHG"="Whiting",
"WHB"="Blue whiting",
"HKE"="Hake",
"MON"="Monkfish",
"HOM"="Horse mackerel",
"MAC"="Mackerel",
"PLE"="Plaice",
"MEG"="Megrim",
"SOL"="Sole",
"BOC"="Boarfish",
"LYY"="Dragonet",
"GUR"="Red gurnard"
)

FDI_EU <- FDI_EU %>%
  mutate(
    species = recode(species, !!!new_species_names)
  )

#figure out which categories of species/gear/country/target assemblage fishing is composed of essentially only discards (no landings)

all_discards<-subset(FDI_EU, discard_rate>1e2)

#remove columns from the discard rate
biomass_discard_rates<-FDI_EU%>%
    filter(!is.na(discard_rate),
           !is.infinite(discard_rate))

```
