---
title: "North Sea Catch Data"
author: "James EV Rimmer"
date: "2025-07-15"
output:
  html_document: default
  pdf_document: default
---

# 1. Introduction

This script processes the landings STECF FDI length-landings data to give landings by gear and size for each species, 2013 - 2021. The script is desgined to flexibly allow the user to modify the subarea selection and the gear classifications.

A data frame is created containing: species (+ 3 letter code), length (in cm), catch (number of individuals), and gear type (e.g., otter, beam, static, pelagic). 

## 2. Data - inputs/outputs
### 2.1 Inputs
data_raw/downloaded - FDI Landings Length.csv

data_raw/hosted - ASFIS_sp_2022.csv

### 2.2 Data description.

Information about the data set can be found here: https://datacollection.jrc.ec.europa.eu/documents/d/dcf/fdi_datacall_annex_2024
Table F contains the column name defintions:

Table F NAO OFR. Landings length data 
Table to be provided only for the Atlantic Ocean and other regions (not for the Mediterranean and Black Sea 
regions). 

1. COUNTRY: According to the code list provided in Appendix 1; missing values not allowed. 
2. YEAR: Four digits (e.g. 2019); missing values not allowed. 
3. DOMAIN_LANDINGS: Text in format specified in Appendix 8; missing values not allowed. 
4. NEP_SUB_REGION: Functional Units (FUs) according to the list in Appendix 9; ‘NK’ if not known or 
‘NA’ if not applicable. This variable is asked only for Nephrops norvegicus catches. 
5. SPECIES: according to the FAO 3-alpha code list, see Appendix 12; missing values not allowed. 
6. TOTWGHTLANDG: Estimated landings in tonnes [precision to 3 digits after the decimal]; missing 
values not allowed. 
7. TOTAL_SAMPLED_TRIPS: The total number of sampled trips that relate to domain; a number 
should only be given only if it relates to this domain, otherwise use ‘NK’. 
8. NO_LENGTH_MEASUREMENTS: The number of length measurements, from within the domain, 
that relate to landings; a number should be given only if it relates to this domain, otherwise ‘NK’. 
9. LENGTH_UNIT: The unit of length classes [‘mm’=millimetre, ‘cm’=centimetre]; ‘NK’ if length 
information is not available. 
10. MIN_LENGTH: [integer] The minimum length in the data for this species-domain combination; 
minimum length and maximum length must either both be ‘NK’ or both be not ‘NK’. 
11. MAX_LENGTH: [integer] The true maximum length in the data for this species-domain 
combination; minimum length and maximum length must either both be ‘NK’ or both be not ‘NK’. 
12. LENGTH: [integer] (MIN_LENGTH <= LENGTH <= MAX_LENGTH). ‘NK’ can be used only if both 
MIN_LENGTH and MAX_LENGTH are not known. 
13. NO_LENGTH: The number in thousands [precision to 3 digits after the decimal] at the specified 
length. If length information is not available, use ‘NK’. 
14. MEAN_WEIGHT_AT_LENGTH: The mean weight of fish; if no age specific information is available 
use ‘NK’. 
15. WEIGHT_UNIT: The unit of measure used for the column MEAN_WEIGHT_AT_LENGTH [kg = 
kilogram, g = gram]; if weight information is not available, use ‘NK’.

Point 13 is relevant, and the units are transformed by a multiplying by 1000. 

### 3. Processing
## 3.1. Importing and curating the data 

```{r, message = F}
## 0. Set up the environment
rm(list = ls())

# Load required libraries
library(tidyverse)
library(here)
library(mizer)
library(mizerExperimental)

FDI_Landings_Length<-read.csv("/Users/jessicawestworth/Desktop/FDI Landings Length.csv")
species_codes <- read.csv("/Users/jessicawestworth/Desktop/ASFIS_sp_2022.csv")

```

## 3.2 Configurable parameters - subareas and gears

```{r}
# Subarea definitions:
#sub_area_7 <- c('27.7.B', '27.7.C', '27.7.E', '27.7.F', '27.7.G', '27.7.H', '27.7.J', '27.7.K')
#ns_sub_area <- c('27.4.A', '27.4.B', '27.4.C')

# Configurable gear groupings:
gear_categories_1 <- list(
  beam = c('TBB', 'PTB'),
  otter = c('OTB', 'OTT', 'OTM'),
  pelagic = c('PTM', 'PS'),
  static = c('GNS', 'GND', 'GNC', 'GTR', 'GTN', 'LLS'),
  other = c('DRB', 'DRH', 'FPO', 'FYK', 'GEF', 'HMD', 'LHM', 'LHP', 'LLD', 'LNB', 'LTL', 'NK', 'NO', 'SB', 'SDN', 'SPR', 'SSC', 'SV')
)

# Create new sub areas of gear categories as required below here:

# sub_area_x <- c()
# gear_categories_2 <- list()

```

## 3.3 Data preparation

```{r}
library(dplyr)
library(stringr)
library(purrr)

# This function now extracts gear types *only* from the 4th underscore-separated domain part
extract_gear_types_combined <- function(domain_parts) {
  if (length(domain_parts) < 4) return(NA_character_)
  gear_block <- domain_parts[4]
  gear_codes <- str_split(gear_block, "-", simplify = FALSE)[[1]]
  paste(gear_codes, collapse = "-")
}

length_landings <- FDI_Landings_Length %>%
  rowwise() %>%
  mutate(
    domain_parts = list(str_split(domain_landings, "_")[[1]]),
    region_code = domain_parts[3],  # e.g. 27.4.a-27.4.b
    gear_type = extract_gear_types_combined(domain_parts),
    
    # Normalize lengths
    length = ifelse(length_unit == "mm", length / 10, length),
    length_unit = "cm",
    source = "landings"
  ) %>%
  ungroup() %>%
  select(-domain_parts) %>%
  left_join(species_codes %>% 
              select(X3A_CODE, Scientific_name, English_name),
            by = c("species" = "X3A_CODE")) %>%
  rename(domain = domain_landings) %>%
  select(
    country, year, domain, species, Scientific_name, English_name,
    length_unit, region_code, gear_type,
    no_length_measurements, length, no_length, mean_weight_at_length,
    source
  )

```

## 3.4 Subarea filtering

```{r}
#Identify any region_code with 27.4... written as sub_region = ns_sub_area
length_landings <- length_landings %>%
  mutate(
    sub_region = case_when(
      str_detect(region_code, "\\b27\\.4\\b") ~ "ns_sub_area",     # exact match
      str_detect(region_code, "27\\.4\\.") ~ "ns_sub_area",        # matches 27.4.a, etc.
      str_detect(region_code, "27\\.4[^0-9]?") ~ "ns_sub_area",    # more general safety net
      TRUE ~ NA_character_
    )
  )

sub_region_landings<-subset(length_landings, sub_region=="ns_sub_area")
```

## 3.5 Calculate catch by species, size, and gear type

```{r}
# Define a function to calculate catch by species, size class, and gear type
catch_by_size_class_and_gear <- function(data) {
    data %>%
        group_by(species, length, gear_type) %>%
        summarise(total_catch = sum(no_length)) %>%
        ungroup()
}

catch_data <- catch_by_size_class_and_gear(sub_region_landings)
# Only if wider-area species required
# catch_data_full_area <- catch_by_size_class_and_gear(full_area_landings)
# catch_data_combined <- bind_rows(catch_data, catch_data_full_area)

species_reference <- length_landings %>%
    distinct(species, Scientific_name, English_name)
```

## 3.6 Add species names and map gear categories

```{r}
catch_data <- catch_data %>%
  left_join(species_reference, by = "species") %>%
  rename(
    gear = gear_type,  # Adjust as needed if `gear_type` is used elsewhere
    catch = total_catch
  ) %>%
  drop_na()  # Remove rows with NA in any column

```

# 4. Generating a size-distributed catch data frame
## 4.1 Create the total gear rows

```{r}
total_gear <- catch_data %>%
  group_by(species, length, Scientific_name, English_name) %>%
  summarise(gear = "total", catch = sum(catch, na.rm = TRUE)) %>%
  ungroup()

combined_data <- bind_rows(catch_data, total_gear)

# Change  "unknown" values to "other"
combined_data <- combined_data %>%
  mutate(gear = ifelse(gear == "unknown", "other", gear))

combined_data$gear <- as.factor(combined_data$gear)

         
         
```

## 4.2 Species filter

```{r}
subset_species <- function(data, species_list, match_by = "species") {
  # Validate match_by parameter
  if (!match_by %in% c("species", "Scientific_name", "English_name")) {
    stop("match_by must be one of 'species', 'Scientific_name', or 'English_name'")
  }
  
  # Filter the data to include only the specified species
  subset_data <- data %>% filter(!!sym(match_by) %in% species_list)
  
  return(subset_data)
}

species_list <- c("SPR", "SAN", "NOP", "HER", "DAB", "WHG", "SOL", "GUX", "GUR", "TUB", "PLE", "HAD", "COD", "POK")

ns_catch <- subset_species(
  data = combined_data,
  species_list = species_list,
  match_by = "species"
 )

# Rename the species column to species_code
colnames(ns_catch)[colnames(ns_catch) == "species"] <- "species_code"

# Create a named vector for the new species names
new_species_names <- c(
  "European sprat" = "Sprat",
  "Sandeel" = "Sandeel",
  "Norway pout" = "N.pout",
  "Atlantic herring" = "Herring",
  "Common dab" = "Dab",
  "Whiting" = "Whiting",
  "Common sole" = "Sole",
  "Gurnards, searobins nei" = "Grey Gurnard",
  "Red gurnard" = "Red Gurnard",
  "Tub Gurnard" = "Tub Gurnard",
  "European plaice" = "Plaice",
  "Haddock" = "Haddock",
  "Atlantic cod" = "Cod",
  "Saithe(=Pollock)" = "Saithe"
)

# Add the new species column by matching English_name
ns_catch$species <- new_species_names[ns_catch$English_name]

# Combine the gurnards
#ns_catch1 <- ns_catch %>%
  # First, standardise species: Grey gurnard or Red gurnard => Gurnard
#  mutate(species = ifelse(species %in% c("Grey Gurnard", "Red Gurnard"), 
 #                         "Gurnard", 
 #                         species)) %>%
  # Now group by length, gear, and the updated species, then sum the catches
#  group_by(length, gear, species) %>%
#  summarise(catch = sum(catch), .groups = "drop")

# Add the dl column
ns_catch$dl <- 1

#saveRDS(ns_catch,"/Users/jessicawestworth/Desktop/Mizer Work/Projects/mizerEcopathedits/inst/extdata/ns_catch.rds")
```








