---
title: "DATRAS Processing survey"
format: pdf
editor: visual
---

Note this seems like a waste of time because isn't the point of matching the survey data also to understand the fishing pressure on fish, survey data does not represent that...

## Load in libraries

```{r}
library(purrr)
library(icesDatras)
library(sf)
library(dplyr)
library(stringr)
library(here)
library(readr)
library(ggplot2)
```

## Load in Surveys

Takes more than an hour to load min to load in data.

```{r}
#SCOROC
SCOROC_HL <- getDATRAS(record = "HL", survey = "SCOROC", year = 2012:2024, quarter = 1:4) 

#BTS
BTS_HL <- getDATRAS(record = "HL", survey = "BTS", year = 2012:2024, quarter = 1:4)

#DYFS
DYFS_HL <- getDATRAS(record = "HL", survey = "DYFS", year = 2012:2024, quarter = 1:4) 

#IE-IGFS
IE_IGFS_HL <- getDATRAS(record = "HL", survey = "IE-IGFS", year = 2012:2024, quarter = 1:4) 

#FR-CGFS
FR_CGFS_HL <- getDATRAS(record = "HL", survey = "FR-CGFS", year = 2012:2024, quarter = 1:4) 

#IE-IAMS
IE_IAMS_HL <- getDATRAS(record = "HL", survey = "IE-IAMS", year = 2012:2024, quarter = 1:4) 

#SP-PORC
SP_PORC_HL <- getDATRAS(record = "HL", survey = "SP-PORC", year = 2012:2024, quarter = 1:4)  
 
#EVHOE
EVHOE_HL <- getDATRAS(record = "HL", survey = "EVHOE", year = 2012:2024, quarter = 1:4) 

#NS-IBTS
NS_IBTS_HL <- getDATRAS(record = "HL", survey = "NS-IBTS", year = 2012:2025, quarter = 1:4) 


#CA data
#SCOROC
SCOROC_CA <- getDATRAS(record = "CA", survey = "SCOROC", year = 2012:2024, quarter = 1:4)
SCOROC_CA <- SCOROC_CA %>%
  select(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, AreaCode) %>%
  distinct(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, .keep_all = TRUE)
SCOROC_HL_CA<-left_join(SCOROC_HL,SCOROC_CA, by=c("Survey","Quarter", "Country", "Ship", "Gear","SweepLngt","GearEx","DoorType","StNo","HaulNo","Year","SpecCode"))

#BTS
BTS_CA <- getDATRAS(record = "CA", survey = "BTS", year = 2012:2024, quarter = 1:4)
BTS_CA <- BTS_CA %>%
  select(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, AreaCode) %>%
  distinct(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, .keep_all = TRUE)
BTS_HL_CA<-left_join(BTS_HL,BTS_CA, by=c("Survey","Quarter", "Country", "Ship", "Gear","SweepLngt","GearEx","DoorType","StNo","HaulNo","Year","SpecCode"))

#DYFS
DYFS_CA <- getDATRAS(record = "CA", survey = "DYFS", year = 2012:2024, quarter = 1:4) 
DYFS_CA <- DYFS_CA %>%
  select(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, AreaCode) %>%
  distinct(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, .keep_all = TRUE)
DYFS_HL_CA<-left_join(DYFS_HL,DYFS_CA, by=c("Survey","Quarter", "Country", "Ship", "Gear","SweepLngt","GearEx","DoorType","StNo","HaulNo","Year","SpecCode"))


#IE-IGFS
IE_IGFS_CA <- getDATRAS(record = "CA", survey = "IE-IGFS", year = 2012:2024, quarter = 1:4) 
IE_IGFS_CA <- IE_IGFS_CA %>%
  select(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, AreaCode) %>%
  distinct(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, .keep_all = TRUE)
IE_IGFS_HL_CA<-left_join(IE_IGFS_HL,IE_IGFS_CA, by=c("Survey","Quarter", "Country", "Ship", "Gear","SweepLngt","GearEx","DoorType","StNo","HaulNo","Year","SpecCode"))

#FR-CGFS
FR_CGFS_CA <- getDATRAS(record = "CA", survey = "FR-CGFS", year = 2012:2024, quarter = 1:4) 
FR_CGFS_CA <- FR_CGFS_CA %>%
  select(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, AreaCode) %>%
  distinct(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, .keep_all = TRUE)
FR_CGFS_HL_CA<-left_join(FR_CGFS_HL,FR_CGFS_CA, by=c("Survey","Quarter", "Country", "Ship", "Gear","SweepLngt","GearEx","DoorType","StNo","HaulNo","Year","SpecCode"))

#IE-IAMS
IE_IAMS_CA <- getDATRAS(record = "CA", survey = "IE-IAMS", year = 2012:2024, quarter = 1:4) 
IE_IAMS_CA <- IE_IAMS_CA %>%
  select(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, AreaCode) %>%
  distinct(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, .keep_all = TRUE)
IE_IAMS_HL_CA<-left_join(IE_IAMS_HL,IE_IAMS_CA, by=c("Survey","Quarter", "Country", "Ship", "Gear","SweepLngt","GearEx","DoorType","StNo","HaulNo","Year","SpecCode"))

#SP-PORC
SP_PORC_CA <- getDATRAS(record = "CA", survey = "SP-PORC", year = 2012:2024, quarter = 1:4)  
SP_PORC_CA <- SP_PORC_CA %>%
  select(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, AreaCode) %>%
  distinct(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, .keep_all = TRUE)
SP_PORC_HL_CA<-left_join(SP_PORC_HL,SP_PORC_CA, by=c("Survey","Quarter", "Country", "Ship", "Gear","SweepLngt","GearEx","DoorType","StNo","HaulNo","Year","SpecCode"))


#EVHOE
EVHOE_CA <- getDATRAS(record = "CA", survey = "EVHOE", year = 2012:2024, quarter = 1:4) 
EVHOE_CA <- EVHOE_CA %>%
  select(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, AreaCode) %>%
  distinct(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, .keep_all = TRUE)
EVHOE_HL_CA<-left_join(EVHOE_HL,EVHOE_CA, by=c("Survey","Quarter", "Country", "Ship", "Gear","SweepLngt","GearEx","DoorType","StNo","HaulNo","Year","SpecCode"))


#NS-IBTS
NS_IBTS_CA <- getDATRAS(record = "CA", survey = "NS-IBTS", year = 2012:2024, quarter = 1:4)
NS_IBTS_CA <- NS_IBTS_CA %>%
  select(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, AreaCode) %>%
  distinct(Survey, Quarter, Country, Ship, Gear, SweepLngt, GearEx, DoorType, StNo, HaulNo, Year, SpecCode, .keep_all = TRUE)
NS_IBTS_HL_CA<-left_join(NS_IBTS_HL,NS_IBTS_CA, by=c("Survey","Quarter", "Country", "Ship", "Gear","SweepLngt","GearEx","DoorType","StNo","HaulNo","Year","SpecCode"))
```

# Clean and Organize Data

Columns shared: Record Type, Quarter, Country, Ship, Gear,SweetLngt,GearEx,DoorType,StNo,HaulNo,Year,SpecCode

```{r}
#rbind data together into singular dataset
data<-rbind(SCOROC_HL_CA,BTS_HL_CA,DYFS_HL_CA,IE_IGFS_HL_CA,FR_CGFS_HL_CA,IE_IAMS_HL_CA,SP_PORC_HL_CA,EVHOE_HL_CA,NS_IBTS_HL_CA)

```

## Subset Desired Celtic Sea Species

```{r}
#remove rows that do not contain the North Sea Desired Species
clean_data<-subset(data, SpecCode %in% c(126417,126436,126437,126438,126439,126484,126555,126822,127023,127143,127146,127160,126792,127259,127419))
#SpecCode to scientific name dictionary is detailed below
```

## Subset Desired Celtic Sea Area

In this section some the the Area Codes generated in the data set loaded from DATRAS are given in different formats, additionally the data frames which load in should have a AreaType Column which should correlate with these codes. After extensive exploration of the data it has become evident this is not the case. Therefore, based on the structure of the AreaCode value an ICES area value is assigned. Only ICES statistical rectangle and ICES area codes are analysed all other area codes are given the value NA and later removed from the data set. This was performed due to the fact that other area codes provided did not give a clear indication that the data included was only from the Celtic Sea and not nearby areas.

### Load in and Apply Dictionary Between ICES rectangles and ICES areas

Processing code of overlapping spatial areas is available in inst/extdata.

```{r}
rect_to_area<-read_csv("/Users/jessicawestworth/Desktop/Mizer Work/Projects/mizerEcopathedits/inst/extdata/ICES Rectangle to Area.csv")
colnames(rect_to_area)<-c("AreaCode","ICES_Area")

data_working <- clean_data %>%
    left_join(rect_to_area, by = c("AreaCode"))

data_add_areas <- data_working %>%
    mutate(
        ICES_Area = case_when(
            is.na(ICES_Area) & AreaCode == "VIIa" ~ "27.7.a",
            is.na(ICES_Area) & AreaCode == "VIIb" ~ "27.7.b",
            is.na(ICES_Area) & AreaCode == "VIIc" ~ "27.7.c",
            is.na(ICES_Area) & AreaCode == "VIId" ~ "27.7.d",
            is.na(ICES_Area) & AreaCode == "VIIe" ~ "27.7.e",
            is.na(ICES_Area) & AreaCode == "VIIf" ~ "27.7.f",
            is.na(ICES_Area) & AreaCode == "VIIg" ~ "27.7.g",
            is.na(ICES_Area) & AreaCode == "VIIh" ~ "27.7.h",
            is.na(ICES_Area) & AreaCode == "VIIj" ~ "27.7.j",
            is.na(ICES_Area) & AreaCode == "VIIk" ~ "27.7.k",
            TRUE ~ ICES_Area  # keep original value if not NA or no match
        )
    )
```

## Subset areas

```{r}
cs_subarea<-c('27.7.b', '27.7.c', '27.7.c.1', '27.7.c.2', '27.7.e', '27.7.f', '27.7.g', '27.7.h', '27.7.j','27.7.j.1','27.7.j.2', '27.7.k','27.7.k.1','27.7.k.2')

cs_data<- data_add_areas %>%
  filter(map_lgl(str_split(ICES_Area, "-"), ~ any(.x %in% cs_subarea)))
```

## Dictionaries

### Species Code

Go to: <https://datras.ices.dk/Data_products/qryspec.aspx>

```{r}
cs_data$SpecCode<-as.factor(cs_data$SpecCode)

SpecCode_to_Latinname<-c("126417"="Clupea harengus",
                         "126436"="Gadus morhua",
                         "126437"="Melanogrammus aeglefinus",
                         "126438"="Merlangius merlangus",
                         "126439"="Micromesistius poutassou",
                         "126484"="Merluccius merluccius",
                         "126555"="Lophius piscatorius",
                         "126822"="Trachurus trachurus",
                         "127023"="Scomber scombrus",
                         "127143"="Pleuronectes platessa",
                         "127146"="Lepidorhombus whiffiagonis",
                         "127160"="Solea solea",
                         "126792"="Callionymus lyra",
                         "127259"="Chelidonichthys cuculus",
                         "127419"="Capros aper")

cs_data$Scientific_name <- SpecCode_to_Latinname[cs_data$SpecCode]
```

## Standardize Length Classes

We want to convert the length to cm. So we will divide LngtClass Columns with LngtCode (".") by 10. BTS surveyalso has a value for 1 which should be centimeters.

```{r}
#check that these are the only codes we have
#make the coversion
cs_survey<-cs_data%>%
    mutate(LngtClass = case_when(
        LngtCode == "."~ LngtClass/10,
        LngtCode == "1"~ LngtClass,
    ))
```

## Add column of tonnes/km\^2/year

```{r}
a_b_conversion <- read_csv("/Users/jessicawestworth/Desktop/Mizer Work/Projects/Mizer_FMP_EwE/Data/life_histories.csv")%>%
    distinct(Species,a,b)

species_map_lauria <- data.frame(
    species = c("Herring", "Cod", "Haddock", "Whiting", "Blue whiting", "Hake", 
                "Monkfish", "Horse mackerel", "Mackerel", "Plaice", "Megrim", 
                "Sole",   "Dragonets", "Red Gurnard", "Boarfish"),
    Scientific_name = c("Clupea harengus", "Gadus morhua", "Melanogrammus aeglefinus", 
                        "Merlangius merlangus", "Micromesistius poutassou", 
                        "Merluccius merluccius", "Lophius piscatorius", 
                        "Trachurus trachurus", "Scomber scombrus", 
                        "Pleuronectes platessa", "Lepidorhombus whiffiagonis", 
                        "Solea solea", "Callionymus spp.", 
                        "Chelidonicthys cuculus","Capros aper")
)

cs_survey <- cs_survey %>%
  group_by(Scientific_name, LngtClass, Gear) %>%
  summarise(
    catch = sum(HLNoAtLngt, na.rm = TRUE),  # sum ignoring NAs
    .groups = "drop")%>%  # ungroup the result
  left_join(species_map_lauria, by="Scientific_name")
cs_survey <- cs_survey %>%
  mutate(LngtClass = floor(LngtClass)) %>%           # Round up lengths
  group_by(species, Scientific_name, Gear, LngtClass) %>%  # Group by relevant columns
  summarise(catch = sum(catch, na.rm = TRUE), .groups = "drop")%>%# Sum catch for duplicates
    rename(length=LngtClass,
           gear=Gear)
cs_survey$dl<-1

cs_survey_final<-cs_survey%>%
    left_join(a_b_conversion, by= c("Scientific_name"="Species"))%>%
    mutate(
        weight=a*length^b,
        scaled_biomass= (weight*catch)/6.174997e+11/(max(clean_data$Year)-min(clean_data$Year)+1),
        scaled_catch=catch/6.174997e+11/(max(clean_data$Year)-min(clean_data$Year)+1))%>%
        select(-a, -b)

#add count column
cs_survey_final$count<-cs_survey_final$catch

#write.csv(cs_survey_final, here("inst", "extdata", "cs_survey.csv"))
```

## Aggregating

```{r}
survey <- cs_survey_final %>%
  group_by(Scientific_name, length, gear) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),  # sum ignoring NAs
    .groups = "drop"  # ungroup the result
  )
survey <- survey %>%
  mutate(length = floor(length)) %>%           # Round up lengths
  group_by(Scientific_name, gear, length) %>%  # Group by relevant columns
  summarise(catch = sum(catch, na.rm = TRUE), .groups = "drop")       # Sum catch for duplicates
survey$dl<-1
colnames(survey)<-c("species","gear","length","catch","dl")

total_survey <- survey %>%
  group_by(species, length, dl) %>%
  summarise(
    gear = "total",                          # Create a new 'gear' category
    catch = sum(catch, na.rm = TRUE),       # Sum all catch across gears
    .groups = "drop"
  )
survey_length_distribution<-total_survey

survey_length_distribution<-survey_length_distribution%>%
    mutate(gear="survey")
#use_data(survey_length_distribution, overwrite=TRUE)
```
