---
title: "Processing Celtic Sea Landings"
format: pdf
editor: visual
---

# Introduction

This script processes the landings STECF FDI length-landings data to give landings by gear and size for each species, 2013 - 2021. The script is desgined to flexibly allow the user to modify the subarea selection and the gear classifications.

A data frame is created containing: species (+ 3 letter code), length (in cm), catch (number of individuals), and gear type (e.g., otter, beam, static, pelagic).

# Data-Inputs/Outputs

## Inputs

data_raw/downloaded - FDI Landings Length.csv

data_raw/hosted - ASFIS_sp_2022.csv

## Data description

Information about the data set can be found here:

<https://datacollection.jrc.ec.europa.eu/documents/d/dcf/fdi_datacall_annex_2025> Table F contains the column name defintions:

Table F NAO OFR. Landings length data Table to be provided only for the Atlantic Ocean and other regions (not for the Mediterranean and Black Sea regions).

1.  COUNTRY: According to the code list provided in Appendix 1; missing values not allowed.
2.  YEAR: Four digits (e.g. 2019); missing values not allowed.
3.  DOMAIN_LANDINGS: Text in format specified in Appendix 8; missing values not allowed.
4.  NEP_SUB_REGION: Functional Units (FUs) according to the list in Appendix 9; ‘NK’ if not known or ‘NA’ if not applicable. This variable is asked only for Nephrops norvegicus catches.
5.  SPECIES: according to the FAO 3-alpha code list, see Appendix 12; missing values not allowed.
6.  TOTWGHTLANDG: Estimated landings in tonnes \[precision to 3 digits after the decimal\]; missing values not allowed.
7.  TOTAL_SAMPLED_TRIPS: The total number of sampled trips that relate to domain; a number should only be given only if it relates to this domain, otherwise use ‘NK’.
8.  NO_LENGTH_MEASUREMENTS: The number of length measurements, from within the domain, that relate to landings; a number should be given only if it relates to this domain, otherwise ‘NK’.
9.  LENGTH_UNIT: The unit of length classes \[‘mm’=millimetre, ‘cm’=centimetre\]; ‘NK’ if length information is not available.
10. MIN_LENGTH: \[integer\] The minimum length in the data for this species-domain combination; minimum length and maximum length must either both be ‘NK’ or both be not ‘NK’.
11. MAX_LENGTH: \[integer\] The true maximum length in the data for this species-domain combination; minimum length and maximum length must either both be ‘NK’ or both be not ‘NK’.
12. LENGTH: \[integer\] (MIN_LENGTH \<= LENGTH \<= MAX_LENGTH). ‘NK’ can be used only if both MIN_LENGTH and MAX_LENGTH are not known.
13. NO_LENGTH: The number in thousands \[precision to 3 digits after the decimal\] at the specified length. If length information is not available, use ‘NK’.
14. MEAN_WEIGHT_AT_LENGTH: The mean weight of fish; if no age specific information is available use ‘NK’.
15. WEIGHT_UNIT: The unit of measure used for the column MEAN_WEIGHT_AT_LENGTH \[kg = kilogram, g = gram\]; if weight information is not available, use ‘NK’.

**Point 13 is relevant, and the units are transformed by a multiplying by 1000.**

## Domain Landings

The domain value of each row contains a variety of information about location and fishing gear. Each category of data supplied is separated by an underscore "\_". If there are multiple entries within a field (e.g. multiple sub-regions, multiple gear types), connect the codes by a dash "-".

The information supplied should follow the structure:

country_quarter\_ region_gear type_target assemblage_mesh size range_selective device_mesh size range of the selective device_vessel length_species_commercial category

There is, however, and exception in which Spain does not supply the country portion. Therefore to allow for Spain's data to be processed properly we need to add the country portion to the domain so that it can be processed properly in the following code.

# Processing

## Importing and curating the data

```{r}
# Load required libraries
library(tidyverse)
library(here)
library(mizer)
library(here)
library(mizerExperimental)

#Need to check if we can upload these to the reprository, if so, we can make this generalised and take from the reprository rather than my machine.
FDI_Landings_Length<-read.csv("/Users/jessicawestworth/Desktop/FDI Landings Length.csv")
FDI_Discards_Length<-read.csv("/Users/jessicawestworth/Desktop/FDI Discards Length.csv")
species_codes <- read.csv("/Users/jessicawestworth/Desktop/ASFIS_sp_2022.csv")
```

## Clean Data

```{r}
library(dplyr)
library(stringr)
library(purrr)

#Remove rows with value NA from the no length column
FDI_Landings_Length_clean <- FDI_Landings_Length[!is.na(FDI_Landings_Length$no_length), ]
FDI_Discards_Length_clean <- FDI_Discards_Length[!is.na(FDI_Discards_Length$no_length), ]


#Multiply no length column by 1000 as the current value is number in thousands
FDI_Landings_Length_clean$no_length<-FDI_Landings_Length_clean$no_length*1000
FDI_Discards_Length_clean$no_length<-FDI_Discards_Length_clean$no_length*1000

#Adding country to the domain value of Spain
FDI_Landings_Length_clean <- FDI_Landings_Length_clean %>%
  mutate(
    domain_landings = if_else(
      country == "Spain" & !startsWith(domain_landings, "ESP_"),
      paste0("ESP_", domain_landings),
      domain_landings
    )
  )

#Remove NIR
FDI_Landings_Length_clean <- FDI_Landings_Length_clean %>%
  filter(!str_starts(domain_landings, "NIR"))

# For discards
FDI_Landings_Length_clean <- FDI_Landings_Length_clean %>%
  mutate(domain_landings = domain_landings %>%
             str_replace_all("_EU_", "_")%>%
             str_replace_all("EU_", "_")%>%
             str_replace_all("_COAST_", "_")%>%
             str_replace_all("COAST_", "_")%>%
             str_replace_all("_UK_", "_"))

FDI_Discards_Length_clean <- FDI_Discards_Length_clean %>%
  mutate(
    domain_discards = if_else(
      country == "Spain" & !startsWith(domain_discards, "ESP_"),
      paste0("ESP_", domain_discards),
      domain_discards
    )
  )

#Remove NIR
FDI_Discards_Length_clean <- FDI_Discards_Length_clean %>%
  filter(!str_starts(domain_discards, "NIR"))

# For discards
FDI_Discards_Length_clean <- FDI_Discards_Length_clean %>%
  mutate(domain_discards = domain_discards %>%
             str_replace_all("_EU_", "_")%>%
             str_replace_all("EU_", "_")%>%
             str_replace_all("_COAST_", "_")%>%
             str_replace_all("COAST_", "_")%>%
             str_replace_all("_UK_", "_"))
```

## Configurable parameters - subareas and gears

The following script is used for the Celtic Sea, however, by defining different subareas one could extract data for a variety of other areas.

```{r}
sub_area_7<- c('27.7.B', '27.7.C', '27.7.E', '27.7.F', '27.7.G', '27.7.H', '27.7.J', '27.7.K')
ns_sub_area<-c('27.4.A','27.4.B','27.4.C') #potentially add 27.4 to this list?
```

## Data preparation

```{r}
# This function extracts gear types *only* from the 4th underscore-separated domain part, ensure that the domains follow the pattern till gear type detailed above before running.
extract_gear_types_combined <- function(domain_parts) {
  if (length(domain_parts) < 4) return(NA_character_)
  gear_block <- domain_parts[4]
  gear_codes <- str_split(gear_block, "-", simplify = FALSE)[[1]]
  paste(gear_codes, collapse = "-")
}

extract_target_assemblage_combined <- function(domain_parts) {
  if (length(domain_parts) < 5) return(NA_character_)
  gear_block <- domain_parts[5]
  gear_codes <- str_split(gear_block, "-", simplify = FALSE)[[1]]
  paste(gear_codes, collapse = "-")
}

length_landings <- FDI_Landings_Length_clean %>%
  rowwise() %>%
  mutate(
    domain_parts = list(str_split(domain_landings, "_")[[1]]),
    region_code = domain_parts[3],  # e.g. 27.4.a-27.4.b
    gear_type = extract_gear_types_combined(domain_parts),
    target_assemblage= extract_target_assemblage_combined(domain_parts),
    # Normalize lengths
    length = ifelse(length_unit == "mm", length / 10, length),
    length_unit = "cm",
    source = "landings"
  ) %>%
  ungroup() %>%
  select(-domain_parts) %>%
  left_join(species_codes %>% 
              select(X3A_CODE, Scientific_name, English_name),
            by = c("species" = "X3A_CODE")) %>%
  rename(domain = domain_landings) %>%
  select(
    country, year, domain, species, Scientific_name, English_name,
    length_unit, region_code, gear_type, target_assemblage,
    no_length_measurements, length, no_length, mean_weight_at_length,
    source
  )

length_discards <- FDI_Discards_Length_clean %>%
  rowwise() %>%
  mutate(
    domain_parts = list(str_split(domain_discards, "_")[[1]]),
    region_code = domain_parts[3],  # e.g. 27.4.a-27.4.b
    gear_type = extract_gear_types_combined(domain_parts),
    target_assemblage= extract_target_assemblage_combined(domain_parts),
    
    # Normalize lengths
    length = ifelse(length_unit == "mm", length / 10, length),
    length_unit = "cm",
    source = "discards"
  ) %>%
  ungroup() %>%
  select(-domain_parts) %>%
  left_join(species_codes %>% 
              select(X3A_CODE, Scientific_name, English_name),
            by = c("species" = "X3A_CODE")) %>%
  rename(domain = domain_discards) %>%
  select(
    country, year, domain, species, Scientific_name, English_name,
    length_unit, region_code, gear_type, target_assemblage,
    no_length_measurements, length, no_length, mean_weight_at_length,
    source
  )
```

## Subarea filtering

Use the previously defined regions above in the configurable parameters section and extract them from the region_code column in the landings_length data frame. This will create a new column called sub_region and give them the value cs_sub area if the values contain any combinations of the defined areas, but reject any rows that contain other sub areas in addition to the defined regions.

Example of treatment:

| region_code   | sub_region  |
|---------------|-------------|
| 27.7.b        | cs_sub_area |
| 27.7.b-27.7.h | cs_sub_area |
| 27.7.g-27.7.a | NA          |
| 27.4.a        | ns_sub_area |
| 27.4.b-27.4.c | ns_sub_area |
| 27.4.a-27.7.h | NA          |

```{r}
length_landings <- length_landings %>%
  mutate(
    # Standardize region_code format (uppercase)
    region_code_upper = toupper(region_code),

    # Split region_code into individual codes (handling multi-code entries)
    region_parts = str_split(region_code_upper, "-"),

    # Assign sub_region if ALL codes in a row match one region group
    sub_region = case_when(
      sapply(region_parts, function(rcs) all(rcs %in% sub_area_7)) ~ "cs_sub_area",
      sapply(region_parts, function(rcs) all(rcs %in% ns_sub_area)) ~ "ns_sub_area",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-region_code_upper, -region_parts)  # Clean up helper columns

length_discards <- length_discards %>%
  mutate(
    # Standardize region_code format (uppercase)
    region_code_upper = toupper(region_code),

    # Split region_code into individual codes (handling multi-code entries)
    region_parts = str_split(region_code_upper, "-"),

    # Assign sub_region if ALL codes in a row match one region group
    sub_region = case_when(
      sapply(region_parts, function(rcs) all(rcs %in% sub_area_7)) ~ "cs_sub_area",
      sapply(region_parts, function(rcs) all(rcs %in% ns_sub_area)) ~ "ns_sub_area",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-region_code_upper, -region_parts)  # Clean up helper columns
```

# Celtic Sea

### Subset data for Celtic Sea

```{r}
cs_sub_region_landings<-subset(length_landings, sub_region=="cs_sub_area")

cs_sub_region_discards<-subset(length_discards, sub_region=="cs_sub_area")
```

### Fixes For Limited Data

In this Celtic Sea example the final data frame from processing has very limited data from the following species: Monkfish and Herring.

Recommended steps:

1\) Check the species filering and cross check that the calls used in species filtering match with the landings_length data frame.

2\) Check the landings_length data frame and determine if there is any landings data from any region for the data limited species.

3\) If data is present for a nearby region document which region you are taking data from for that species and follow the recommended code below. *Reference for checking proximity to desired area: <https://fish-commercial-names.ec.europa.eu/fish-names/fishing-areas/fao-area-27_en>*

4\) If data is not present for that species in a nearby region utilize an alternative data source.

#### Herring

There is data available for the nearby region 27.7.d

```{r}
#subset for herring
herring<-subset(length_landings, species=="HER")

#subset herring at region 27.7.d
herring_27.7<-subset(herring, region_code==c("27.7.d","27.7"))
```

#### Monkfish

There is data available for the region 27.3.a.20

```{r}
#subset for monkfish
monkfish<-subset(length_landings, species=="MON")

#subset monkfish at region 27.3.a.20
monkfish_27.3.a.20<-subset(monkfish, region_code=="27.3.a.20")
```

#### Bind data

```{r}
cs_sub_region_landings<-rbind(cs_sub_region_landings,monkfish_27.3.a.20,herring_27.7)
```

### Calculate catch by species, size, and gear type

```{r}
# Define a function to calculate catch by species, size class, and gear type
catch_by_size_class_and_gear <- function(data) {
    data %>%
        group_by(species, length, gear_type, target_assemblage, country, source) %>%
        summarise(total_catch = sum(no_length)) %>%
        ungroup()
}

cs_catch_data <- catch_by_size_class_and_gear(cs_sub_region_landings)

species_reference_landings <- length_landings %>%
    distinct(species, Scientific_name, English_name)

cs_discards_data <- catch_by_size_class_and_gear(cs_sub_region_discards)

species_reference_discards <- length_discards %>%
    distinct(species, Scientific_name, English_name)
```

### Add species names and map gear categories

```{r}
cs_catch_data <- cs_catch_data %>%
  left_join(species_reference_landings, by = "species") %>%
  rename(
    gear = gear_type,  # Adjust as needed if `gear_type` is used elsewhere
    catch = total_catch,
    target_assemblage =target_assemblage,
    country=country
  ) %>%
  drop_na()  # Remove rows with NA in any column

cs_discards_data <- cs_discards_data %>%
  left_join(species_reference_discards, by = "species") %>%
  rename(
    gear = gear_type,  # Adjust as needed if `gear_type` is used elsewhere
    catch = total_catch,
    target_assemblage =target_assemblage,
    country=country
  ) %>%
  drop_na()  # Remove rows with NA in any column
```

## Generating a size-distributed catch data frame

### Species filter

cs_discards

```{r}
subset_species <- function(data, species_list, match_by = "species") {
  # Validate match_by parameter
  if (!match_by %in% c("species", "Scientific_name", "English_name")) {
    stop("match_by must be one of 'species', 'Scientific_name', or 'English_name'")
  }
  
  # Filter the data to include only the specified species
  subset_data <- data %>% filter(!!sym(match_by) %in% species_list)
  
  return(subset_data)
}

species_list <- c("HER","COD","HAD","WHG","WHB","HKE","MON","HOM","MAC","PLE","MEG","SOL","GUR", "BOC", "LYY")

cs_landings <- subset_species(
  data = cs_catch_data,
  species_list = species_list,
  match_by = "species"
 )

cs_discards <- subset_species(
  data = cs_discards_data,
  species_list = species_list,
  match_by = "species"
 )

# Rename the species column to species_code
colnames(cs_landings)[colnames(cs_landings) == "species"] <- "species_code"

colnames(cs_discards)[colnames(cs_discards) == "species"] <- "species_code"

# Create a named vector for the new species names
new_species_names <- c("Atlantic herring"="Herring",
"Atlantic cod"="Cod", 
"Haddock"="Haddock",
"Whiting"="Whiting",
"Blue whiting(=Poutassou)"="Blue whiting",
"European hake"="Hake",
"Angler(=Monk)"="Monkfish",
"Atlantic horse mackerel"="Horse mackerel",
"Atlantic mackerel"="Mackerel",
"European plaice"="Plaice",
"Megrim"="Megrim",
"Common sole"="Sole",
"Boarfish"="Boarfish",
"Dragonet"="Dragonet",
"Red gurnard"="Red gurnard"
)

# Add the new species column by matching English_name
cs_landings$species <- new_species_names[cs_landings$English_name]

cs_discards$species <- new_species_names[cs_discards$English_name]

# Add the dl column
cs_landings$dl <- 1
cs_discards$dl <- 1

#convert catch to per year per area (m^2): total area surveyed is area of celtic sea 6.174997e+11 (m^2) and number of years is 2013 to 2020 so 8 years
#cs_landings$catch<-cs_landings$catch/6.174997e+11/8
#saveRDS(cs_catch, here("inst","extdata","cs_catch.rds"))
#cs_discards$catch<-cs_discards$catch/6.174997e+11/8
```

Combining gears to ecopath gears

```{r}
#Discard rate for species from Demersal Atlas for discards
cs_landings <- cs_landings %>%
  mutate(
    gear_category = case_when(
      gear %in% c("OTT","OTB","OTB-OTT-PTB","TBB") ~ "Trawl",
      gear %in% c("GND-GNS-GNC","GNS","GTR","GTR-GTN") ~ "Gillnet",
      gear %in% c("LHM-LHP","OTM","OTM-PTM","LLS","PS","PS-LA","SSC") ~ "Pelagic",
      TRUE ~ "Misc") # default for everything else
  )

cs_discards <- cs_discards %>%
  mutate(
    gear_category = case_when(
      gear %in% c("OTT","OTB","OTB-OTT-PTB","TBB") ~ "Trawl",
      gear %in% c("GND-GNS-GNC","GNS","GTR","GTR-GTN") ~ "Gillnet",
      gear %in% c("LHM-LHP","OTM","OTM-PTM","LLS","PS","PS-LA","SSC") ~ "Pelagic",
      TRUE ~ "Misc") # default for everything else
  )

cs_landings <- cs_landings %>%
  group_by(species, Scientific_name, length,gear_category, target_assemblage, source) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),       # Sum all catch across gears
    .groups = "drop"
  )

cs_discards <- cs_discards %>%
  group_by(species, Scientific_name, length,gear_category, target_assemblage, source) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),       # Sum all catch across gears
    .groups = "drop"
  )
```

Normalise the catches

```{r}

life_history_fishbase <- read.csv(here("inst","extdata","Celtic_Sea_Fishbase_life_History_Params.csv"))

cs_landings<-cs_landings%>%
    left_join(life_history_fishbase, by=c("species","Scientific_name"))%>%
    mutate(weight=a*length^b,
           biomass=weight*catch)

cs_discards<-cs_discards%>%
    left_join(life_history_fishbase, by=c("species","Scientific_name"))%>%
    mutate(weight=a*length^b,
           biomass=weight*catch)

cs_landings_normalised <- cs_landings %>%
  group_by(species, Scientific_name, gear_category,target_assemblage) %>%         
  mutate(
    total_catch = sum(catch, na.rm = TRUE),
    total_biomass = sum(biomass, na.rm = TRUE)) %>%
  ungroup()%>%  
    mutate(
    catch_proportion = catch / total_catch,
    biomass_proportion = biomass / total_biomass)%>%
    select(-total_catch, -catch, -source, -total_biomass, -biomass, -a,-b,-l_mat,-l_max)

cs_discards_normalised<- cs_discards %>%
  group_by(species, Scientific_name, gear_category, target_assemblage) %>%         
  mutate(
    total_catch = sum(catch, na.rm = TRUE),
    total_biomass = sum(biomass, na.rm = TRUE)) %>%
  ungroup()%>%  
    mutate(
    catch_proportion = catch / total_catch,
    biomass_proportion = biomass / total_biomass)%>%
    select(-total_catch, -catch, -source, -total_biomass, -biomass, -a,-b,-l_mat,-l_max)
```

Combining discards and landings

```{r}
#Blue Whiting
blue_whiting_cs_landings<-subset(cs_landings_normalised, species=="Blue whiting")
blue_whiting_cs_discards<-subset(cs_discards_normalised, species=="Blue whiting")
blue_whiting_cs_combined<-blue_whiting_cs_landings%>%
    full_join(blue_whiting_cs_discards, by=c("species","Scientific_name",
                                             "length", "weight",
                                             "gear_category","target_assemblage"))%>%
    mutate(biomass_proportion.x = (replace_na(catch_proportion.x, 0))*(1-0.261),
           biomass_proportion.y = (replace_na(catch_proportion.y, 0))*0.261,
           catch_proportion.x = biomass_proportion.x/weight,
           catch_proportion.y = (biomass_proportion.y/weight)*0.5,
           catch = catch_proportion.x + catch_proportion.y)%>%
    select(-biomass_proportion.x,-biomass_proportion.y,-catch_proportion.x,
           -biomass_proportion.y,-target_assemblage)

blue_whiting_cs <- blue_whiting_cs_combined %>%
  group_by(species, Scientific_name, gear_category, length) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name, gear_category) %>%
    mutate(total_catch = sum(catch, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch / total_catch)%>%
    select(-total_catch, -catch)

#Boarfish
boarfish_cs_landings<-subset(cs_landings_normalised, species=="Boarfish")
boarfish_cs_discards<-subset(cs_discards_normalised, species=="Boarfish")
#only discards for boarfish, can be added to the main data frame w/o scaling
boarfish_cs_combined<-boarfish_cs_landings%>%
    full_join(boarfish_cs_discards, by=c("species","Scientific_name",
                                             "length", "weight",
                                             "gear_category","target_assemblage"))%>%
    mutate(biomass_proportion.x = (replace_na(catch_proportion.x, 0)),
           biomass_proportion.y = (replace_na(catch_proportion.y, 0)),
           catch_proportion.x = biomass_proportion.x/weight,
           catch_proportion.y = (biomass_proportion.y/weight)*0.5,
           catch = catch_proportion.x + catch_proportion.y)%>%
    select(-biomass_proportion.x,-biomass_proportion.y,-catch_proportion.x,
           -biomass_proportion.y,-target_assemblage)

boarfish_cs <- boarfish_cs_combined %>%
  group_by(species, Scientific_name, gear_category,length) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name, gear_category) %>%
    mutate(total_catch = sum(catch, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch / total_catch)%>%
    select(-total_catch, -catch)

#Cod
cod_cs_landings<-subset(cs_landings_normalised, species=="Cod")
cod_cs_discards<-subset(cs_discards_normalised, species=="Cod")
#assume discards for demersal fishing for cod is 11% and we assume a 50% survival rate
cod_cs_combined<-cod_cs_landings%>%
    full_join(cod_cs_discards, by=c("species","Scientific_name",
                                             "length", "weight",
                                             "gear_category","target_assemblage"))%>%
    mutate(biomass_proportion.x = (replace_na(catch_proportion.x, 0))*(1-0.11),
           biomass_proportion.y = (replace_na(catch_proportion.y, 0))*0.11,
           catch_proportion.x = biomass_proportion.x/weight,
           catch_proportion.y = (biomass_proportion.y/weight)*0.5,
           catch = catch_proportion.x + catch_proportion.y)%>%
    select(-biomass_proportion.x,-biomass_proportion.y,-catch_proportion.x,
           -biomass_proportion.y,-target_assemblage)

cod_cs <- cod_cs_combined %>%
  group_by(species, Scientific_name, gear_category, length) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name, gear_category) %>%
    mutate(total_catch = sum(catch, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch / total_catch)%>%
    select(-total_catch, -catch)

#Dragonet
dragonet_cs_landings<-subset(cs_landings_normalised, species=="Dragonet")
dragonet_cs_discards<-subset(cs_discards_normalised, species=="Dragonet")

dragonet_cs_combined<-dragonet_cs_landings%>%
    full_join(dragonet_cs_discards, by=c("species","Scientific_name",
                                             "length", "weight",
                                             "gear_category","target_assemblage"))%>%
    mutate(biomass_proportion.x = (replace_na(catch_proportion.x, 0)),
           biomass_proportion.y = (replace_na(catch_proportion.y, 0)),
           catch_proportion.x = biomass_proportion.x/weight,
           catch_proportion.y = (biomass_proportion.y/weight)*0.5,
           catch = catch_proportion.x + catch_proportion.y)%>%
    select(-biomass_proportion.x,-biomass_proportion.y,-catch_proportion.x,
           -biomass_proportion.y,-target_assemblage)

dragonet_cs <- dragonet_cs_combined %>%
  group_by(species, Scientific_name,gear_category,length) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name,gear_category) %>%
    mutate(total_catch = sum(catch, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch / total_catch)%>%
    select(-total_catch, -catch)

#Haddock
haddock_cs_landings<-subset(cs_landings_normalised, species=="Haddock")
haddock_cs_discards<-subset(cs_discards_normalised, species=="Haddock")

haddock_cs_combined<-haddock_cs_landings%>%
    full_join(haddock_cs_discards, by=c("species","Scientific_name",
                                             "length", "weight",
                                             "gear_category","target_assemblage"))%>%
    mutate(biomass_proportion.x = (replace_na(catch_proportion.x, 0))*(1-0.55),
           biomass_proportion.y = (replace_na(catch_proportion.y, 0))*0.55,
           catch_proportion.x = biomass_proportion.x/weight,
           catch_proportion.y = (biomass_proportion.y/weight)*0.5,
           catch = catch_proportion.x + catch_proportion.y)%>%
    select(-biomass_proportion.x,-biomass_proportion.y,-catch_proportion.x,
           -biomass_proportion.y,-target_assemblage)

haddock_cs <- haddock_cs_combined %>%
  group_by(species, Scientific_name, gear_category,length) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name, gear_category) %>%
    mutate(total_catch = sum(catch, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch / total_catch)%>%
    select(-total_catch, -catch)

#Hake
hake_cs_landings<-subset(cs_landings_normalised, species=="Hake")
hake_cs_discards<-subset(cs_discards_normalised, species=="Hake")
hake_cs_combined<-hake_cs_landings%>%
    full_join(hake_cs_discards, by=c("species","Scientific_name",
                                             "length", "weight",
                                             "gear_category","target_assemblage"))%>%
    mutate(biomass_proportion.x = (replace_na(catch_proportion.x, 0))*(1-0.34),
           biomass_proportion.y = (replace_na(catch_proportion.y, 0))*0.34,
           catch_proportion.x = biomass_proportion.x/weight,
           catch_proportion.y = (biomass_proportion.y/weight)*0.5,
           catch = catch_proportion.x + catch_proportion.y)%>%
    select(-biomass_proportion.x,-biomass_proportion.y,-catch_proportion.x,
           -biomass_proportion.y,-target_assemblage)
hake_cs <- hake_cs_combined %>%
  group_by(species, Scientific_name, gear_category,length) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name, gear_category) %>%
    mutate(total_catch = sum(catch, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch / total_catch)%>%
    select(-total_catch, -catch)

#Herring
herring_cs_landings<-subset(cs_landings_normalised, species=="Herring")
herring_cs_discards<-subset(cs_discards_normalised, species=="Herring")

herring_cs_combined<-herring_cs_landings%>%
    full_join(herring_cs_discards, by=c("species","Scientific_name",
                                             "length", "weight",
                                             "gear_category","target_assemblage"))%>%
    mutate(biomass_proportion.x = (replace_na(catch_proportion.x, 0))*(1-0.01),
           biomass_proportion.y = (replace_na(catch_proportion.y, 0))*0.01,
           catch_proportion.x = biomass_proportion.x/weight,
           catch_proportion.y = (biomass_proportion.y/weight)*0.5,
           catch = catch_proportion.x + catch_proportion.y)%>%
    select(-biomass_proportion.x,-biomass_proportion.y,-catch_proportion.x,
           -biomass_proportion.y,-target_assemblage)

herring_cs <- herring_cs_combined %>%
  group_by(species, Scientific_name, gear_category,length) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name, gear_category) %>%
    mutate(total_catch = sum(catch, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch / total_catch)%>%
    select(-total_catch, -catch)

#Horse mackerel
horse_mackerel_cs_landings<-subset(cs_landings_normalised, species=="Horse mackerel")
horse_mackerel_cs_discards<-subset(cs_discards_normalised, species=="Horse mackerel")

horse_mackerel_cs_combined<-horse_mackerel_cs_landings%>%
    full_join(horse_mackerel_cs_discards, by=c("species","Scientific_name",
                                             "length", "weight",
                                             "gear_category","target_assemblage"))%>%
    mutate(biomass_proportion.x = (replace_na(catch_proportion.x, 0))*(1-0.05),
           biomass_proportion.y = (replace_na(catch_proportion.y, 0))*0.05,
           catch_proportion.x = biomass_proportion.x/weight,
           catch_proportion.y = (biomass_proportion.y/weight)*0.5,
           catch = catch_proportion.x + catch_proportion.y)%>%
    select(-biomass_proportion.x,-biomass_proportion.y,-catch_proportion.x,
           -biomass_proportion.y,-target_assemblage)

horse_mackerel_cs <- horse_mackerel_cs_combined %>%
  group_by(species, Scientific_name, gear_category,length) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name, gear_category) %>%
    mutate(total_catch = sum(catch, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch / total_catch)%>%
    select(-total_catch, -catch)

#Mackerel
mackerel_cs_landings<-subset(cs_landings_normalised, species=="Mackerel")
mackerel_cs_discards<-subset(cs_discards_normalised, species=="Mackerel")

mackerel_cs_combined<-mackerel_cs_landings%>%
    full_join(mackerel_cs_discards, by=c("species","Scientific_name",
                                             "length", "weight",
                                             "gear_category","target_assemblage"))%>%
    mutate(biomass_proportion.x = (replace_na(catch_proportion.x, 0))*(1-0.01),
           biomass_proportion.y = (replace_na(catch_proportion.y, 0))*0.01,
           catch_proportion.x = biomass_proportion.x/weight,
           catch_proportion.y = (biomass_proportion.y/weight)*0.5,
           catch = catch_proportion.x + catch_proportion.y)%>%
    select(-biomass_proportion.x,-biomass_proportion.y,-catch_proportion.x,
           -biomass_proportion.y,-target_assemblage)

mackerel_cs <- mackerel_cs_combined %>%
  group_by(species, Scientific_name, gear_category,length) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name, gear_category) %>%
    mutate(total_catch = sum(catch, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch / total_catch)%>%
    select(-total_catch, -catch)

#Megrim
megrim_cs_landings<-subset(cs_landings_normalised, species=="Megrim")
megrim_cs_discards<-subset(cs_discards_normalised, species=="Megrim")

megrim_cs_combined<-megrim_cs_landings%>%
    full_join(megrim_cs_discards, by=c("species","Scientific_name",
                                             "length", "weight",
                                             "gear_category","target_assemblage"))%>%
    mutate(biomass_proportion.x = (replace_na(catch_proportion.x, 0))*(1-0.29),
           biomass_proportion.y = (replace_na(catch_proportion.y, 0))*0.29,
           catch_proportion.x = biomass_proportion.x/weight,
           catch_proportion.y = (biomass_proportion.y/weight)*0.5,
           catch = catch_proportion.x + catch_proportion.y)%>%
    select(-biomass_proportion.x,-biomass_proportion.y,-catch_proportion.x,
           -biomass_proportion.y,-target_assemblage)

megrim_cs <- megrim_cs_combined %>%
  group_by(species, Scientific_name, gear_category,length) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name, gear_category) %>%
    mutate(total_catch = sum(catch, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch / total_catch)%>%
    select(-total_catch, -catch)

#Monkfish
monkfish_cs_landings<-subset(cs_landings_normalised, species=="Monkfish")
monkfish_cs_discards<-subset(cs_discards_normalised, species=="Monkfish")

monkfish_cs_combined<-monkfish_cs_landings%>%
    full_join(monkfish_cs_discards, by=c("species","Scientific_name",
                                             "length", "weight",
                                             "gear_category","target_assemblage"))%>%
    mutate(biomass_proportion.x = (replace_na(catch_proportion.x, 0))*(1-0.18),
           biomass_proportion.y = (replace_na(catch_proportion.y, 0))*0.18,
           catch_proportion.x = biomass_proportion.x/weight,
           catch_proportion.y = (biomass_proportion.y/weight)*0.5,
           catch = catch_proportion.x + catch_proportion.y)%>%
    select(-biomass_proportion.x,-biomass_proportion.y,-catch_proportion.x,
           -biomass_proportion.y,-target_assemblage)

monkfish_cs <- monkfish_cs_combined %>%
  group_by(species, Scientific_name, gear_category,length) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name, gear_category) %>%
    mutate(total_catch = sum(catch, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch / total_catch)%>%
    select(-total_catch, -catch)

#Plaice
plaice_cs_landings<-subset(cs_landings_normalised, species=="Plaice")
plaice_cs_discards<-subset(cs_discards_normalised, species=="Plaice")

plaice_cs_combined<-plaice_cs_landings%>%
    full_join(plaice_cs_discards, by=c("species","Scientific_name",
                                             "length", "weight",
                                             "gear_category","target_assemblage"))%>%
    mutate(biomass_proportion.x = (replace_na(catch_proportion.x, 0))*(1-0.71),
           biomass_proportion.y = (replace_na(catch_proportion.y, 0))*0.71,
           catch_proportion.x = biomass_proportion.x/weight,
           catch_proportion.y = (biomass_proportion.y/weight)*0.5,
           catch = catch_proportion.x + catch_proportion.y)%>%
    select(-biomass_proportion.x,-biomass_proportion.y,-catch_proportion.x,
           -biomass_proportion.y,-target_assemblage)

plaice_cs <- plaice_cs_combined %>%
  group_by(species, Scientific_name, gear_category,length) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name,gear_category) %>%
    mutate(total_catch = sum(catch, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch / total_catch)%>%
    select(-total_catch, -catch)

#Red gurnard
red_gurnard_cs_landings<-subset(cs_landings_normalised, species=="Red gurnard")
red_gurnard_cs_discards<-subset(cs_discards_normalised, species=="Red gurnard")

red_gurnard_cs_combined<-red_gurnard_cs_landings%>%
    full_join(red_gurnard_cs_discards, by=c("species","Scientific_name",
                                             "length", "weight",
                                             "gear_category","target_assemblage"))%>%
    mutate(biomass_proportion.x = (replace_na(catch_proportion.x, 0))*(1-0.70),
           biomass_proportion.y = (replace_na(catch_proportion.y, 0))*0.70,
           catch_proportion.x = biomass_proportion.x/weight,
           catch_proportion.y = (biomass_proportion.y/weight)*0.5,
           catch = catch_proportion.x + catch_proportion.y)%>%
    select(-biomass_proportion.x,-biomass_proportion.y,-catch_proportion.x,
           -biomass_proportion.y,-target_assemblage)

red_gurnard_cs <- red_gurnard_cs_combined %>%
  group_by(species, Scientific_name, gear_category,length) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name, gear_category) %>%
    mutate(total_catch = sum(catch, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch / total_catch)%>%
    select(-total_catch, -catch)

#Sole
sole_cs_landings<-subset(cs_landings_normalised, species=="Sole")
sole_cs_discards<-subset(cs_discards_normalised, species=="Sole")

sole_cs_combined<-sole_cs_landings%>%
    full_join(sole_cs_discards, by=c("species","Scientific_name",
                                             "length", "weight",
                                             "gear_category","target_assemblage"))%>%
    mutate(biomass_proportion.x = (replace_na(catch_proportion.x, 0))*(1-0.05),
           biomass_proportion.y = (replace_na(catch_proportion.y, 0))*0.05,
           catch_proportion.x = biomass_proportion.x/weight,
           catch_proportion.y = (biomass_proportion.y/weight)*0.5,
           catch = catch_proportion.x + catch_proportion.y)%>%
    select(-biomass_proportion.x,-biomass_proportion.y,-catch_proportion.x,
           -biomass_proportion.y,-target_assemblage)

sole_cs <- sole_cs_combined %>%
  group_by(species, Scientific_name, gear_category,length) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name, gear_category) %>%
    mutate(total_catch = sum(catch, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch / total_catch)%>%
    select(-total_catch, -catch)

#Whiting
whiting_cs_landings<-subset(cs_landings_normalised, species=="Whiting")
whiting_cs_discards<-subset(cs_discards_normalised, species=="Whiting")

whiting_cs_combined<-whiting_cs_landings%>%
    full_join(whiting_cs_discards, by=c("species","Scientific_name",
                                             "length", "weight",
                                             "gear_category","target_assemblage"))%>%
    mutate(biomass_proportion.x = (replace_na(catch_proportion.x, 0))*(1-0.54),
           biomass_proportion.y = (replace_na(catch_proportion.y, 0))*0.54,
           catch_proportion.x = biomass_proportion.x/weight,
           catch_proportion.y = (biomass_proportion.y/weight)*0.5,
           catch = catch_proportion.x + catch_proportion.y)%>%
    select(-biomass_proportion.x,-biomass_proportion.y,-catch_proportion.x,
           -biomass_proportion.y,-target_assemblage)

whiting_cs <- whiting_cs_combined %>%
  group_by(species, Scientific_name, gear_category,length) %>%
  summarise(
    catch = sum(catch, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name, gear_category) %>%
    mutate(total_catch = sum(catch, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch / total_catch)%>%
    select(-total_catch, -catch)
```

Combining countries

```{r}
Celtic_Sea_Fishing<-rbind(blue_whiting_cs, boarfish_cs,cod_cs, dragonet_cs, haddock_cs,hake_cs,herring_cs,horse_mackerel_cs,mackerel_cs,megrim_cs,monkfish_cs, plaice_cs,red_gurnard_cs,sole_cs,whiting_cs)

Celtic_Sea_Fishing <- Celtic_Sea_Fishing %>%
  group_by(species, Scientific_name, gear_category,length) %>%
  summarise(
    catch_proportion = sum(catch_proportion, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name, gear_category) %>%
    mutate(total_catch = sum(catch_proportion, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    catch_proportion = catch_proportion / total_catch,
    dl= 1)%>%
    select(-total_catch)

Celtic_Sea_Fishing_Mix <- Celtic_Sea_Fishing %>%
    group_by(species, Scientific_name, length) %>%
  summarise(
    catch_proportion = sum(catch_proportion, na.rm = TRUE),
    .groups = "drop")%>%
    group_by(species, Scientific_name) %>%
    mutate(total_catch = sum(catch_proportion, na.rm = TRUE))%>%
    ungroup()%>%
    mutate(
    species=species,
    Scientific_name=Scientific_name,
    gear_category="Mix",
    length=length,
    catch_proportion = catch_proportion / total_catch,
    dl= 1,
    )%>%
    select(-total_catch)

Celtic_Sea_Fishing<-rbind(Celtic_Sea_Fishing,Celtic_Sea_Fishing_Mix)
```

Load in FDI landings data per species and gear and use to get proportion of total landings then multiply by total landings from ICES of the total landed multiply and add the percentage of weight surviving and the percentage survival

```{r}
FDI_EU_2013<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2013.csv")
FDI_EU_2014<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2014.csv")
FDI_EU_2015<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2015.csv")
FDI_EU_2016<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2016.csv")
FDI_EU_2017<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2017.csv")
FDI_EU_2018<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2018.csv")
FDI_EU_2019<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2019.csv")
FDI_EU_2020<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2020.csv")
FDI_EU_2021<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2021.csv")
FDI_EU_2022<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2022.csv")
FDI_EU_2023<-read.csv("/Users/jessicawestworth/Desktop/Data we might want/Catches/FDI Catches by country2023.csv")

FDI_EU<-rbind(FDI_EU_2013,FDI_EU_2014,FDI_EU_2015,FDI_EU_2016,FDI_EU_2017,FDI_EU_2018,FDI_EU_2019,FDI_EU_2020,FDI_EU_2021,FDI_EU_2022,FDI_EU_2023)

unique(FDI_EU$sub_region)

FDI_EU <- FDI_EU %>%
  mutate(
    sub_region = case_when(
      sub_region %in% c("27.7.B") ~ "cs",
      sub_region %in% c("27.7.D") ~ "cs",
      sub_region %in% c("27.7.E") ~ "cs",
      sub_region %in% c("27.7.F") ~ "cs",
      sub_region %in% c("27.7.G") ~ "cs",
      sub_region %in% c("27.7.H") ~ "cs",
      sub_region %in% c("27.7.J") ~ "cs",
      TRUE ~ NA_character_),
    total_live_weight_landed = case_when(
        total_live_weight_landed %in% c("C") ~ NA_character_,
        TRUE ~ total_live_weight_landed),
    tot_discards_tonnes = case_when(
        tot_discards_tonnes %in% c("C") ~ NA_character_,
        TRUE ~ tot_discards_tonnes))

FDI_EU<-subset(FDI_EU, sub_region=="cs")
FDI_EU$tot_discards_tonnes<-as.numeric(FDI_EU$tot_discards_tonnes)
FDI_EU<-FDI_EU[c("gear_type","species","total_live_weight_landed","tot_discards_tonnes")]
FDI_EU$tot_discards_tonnes<-replace_na(FDI_EU$tot_discards_tonnes, 0)
FDI_EU$total_live_weight_landed<-as.numeric(FDI_EU$total_live_weight_landed)
FDI_EU<-na.omit(FDI_EU)

FDI_EU<-FDI_EU %>%
    group_by(gear_type, species)%>%
    summarise(total_landings = sum(total_live_weight_landed),
              tot_discards_tonnes = sum(tot_discards_tonnes))

species_list <- c("HER","COD","HAD","WHG","WHB","HKE","MON","HOM","MAC","PLE","MEG","SOL","GUR", "BOC", "LYY")

FDI_EU <- subset_species(
  data = FDI_EU,
  species_list = species_list,
  match_by = "species"
 )

FDI_EU <- FDI_EU%>%
    mutate(
        gear_category = case_when(
             gear_type %in% c("OTT","OTB","OTB-OTT-PTB","PTB","TBB") ~ "Trawl",
             gear_type %in% c("GND-GNS-GNC","GNS","GTR","GTR-GTN",
                               "GND","GTN","GNC") ~ "Gillnet",
             gear_type %in% c("LNB","LTL","LHM-LHP","LHM","LHP","LLD","OTM","OTM-PTM",
                              "PTM","SB","SDN","SV","SPR","LLS","PS","PS-LA",
                              "SSC","GEF") ~ "Pelagic",
             TRUE ~ "Misc"))%>% # default for everything else
    group_by(gear_category, species)%>%
    summarise(total_landings = sum(total_landings),
              tot_discards_tonnes =sum(tot_discards_tonnes))

#get total f mort proportions
FDI_EU <- FDI_EU%>%
    group_by(gear_category, species)%>%
    mutate(
    total_weight_dead=total_landings+0.5*tot_discards_tonnes)

FDI_EU <- FDI_EU %>%
    group_by(species)%>%
    mutate(
    total_weight_dead_sp=sum(total_weight_dead))%>%
    ungroup()%>%
    mutate(
    proportion_of_total_weight_dead = total_weight_dead / total_weight_dead_sp,
    species_code=species)%>%
    select(-total_weight_dead, -total_weight_dead_sp, -species)

yield <- read.csv(here("inst","extdata","Ecopath-Fishing mortality rates.csv"))
colnames(yield)[2] <- "Group.name"
yield$Total<-rowSums(yield[,3:29],na.rm=TRUE)
yield <- left_join(ecopath_params, yield, by = c("Group.name" = "Group.name")) |>
    mutate(TotalCatch..t.km..year. = Total * Biomass..t.km.2.) |>
    select(Group.name, TotalCatch..t.km..year.)

yield<-yield%>%
    mutate(
        species = case_when(
             Group.name %in% c("Cod","Juvenile cod") ~ "Cod",
             Group.name %in% c("Blue whiting","Juvenile blue whiting") ~ "Blue whiting",
             Group.name %in% c("Hake","Juvenile hake") ~ "Hake",
             Group.name %in% c("Plaice","Juvenile plaice") ~ "Plaice",
             Group.name %in% c("Megrim","Juvenile megrim") ~ "Megrim",
             Group.name %in% c("Whiting","Juvenile whiting") ~ "Whiting",
             Group.name %in% c("Monkfish","Juvenile monkfish") ~ "Monkfish",
             Group.name %in% c("Haddock","Juvenile haddock") ~ "Haddock",
             Group.name %in% c("Sole") ~ "Sole",
             Group.name %in% c("Mackerel") ~ "Mackerel",
             Group.name %in% c("Horse mackerel") ~ "Horse mackerel",
             Group.name %in% c("Herring") ~ "Herring",
             TRUE ~ NA_character_),
        species_code = case_when(
            species %in% c("Cod") ~ "COD",
            species %in% c("Blue whiting") ~ "WHB",
            species %in% c("Hake") ~ "HKE",
            species %in% c("Plaice") ~ "PLE",
            species %in% c("Megrim") ~ "MEG",
            species %in% c("Whiting") ~ "WHG",
            species %in% c("Monkfish") ~ "MON",
            species %in% c("Haddock") ~ "HAD",
            species %in% c("Sole") ~ "SOL",
            species %in% c("Mackerel") ~ "MAC",
            species %in% c("Horse mackerel") ~ "HOM",
            species %in% c("Herring") ~ "HER",
        ))%>%
    group_by(species, species_code)%>%
    summarise(total_weight_dead=sum(TotalCatch..t.km..year., na.rm=TRUE))

yield<-na.omit(yield)  
Red_gurnard<-data.frame(species="Red gurnard",species_code="GUR", total_weight_dead=0.0223)
Boarfish<-data.frame(species="Boarfish",species_code="BOC", total_weight_dead=0.01911)
Dragonet<-data.frame(species="Dragonet",species_code="LYY",total_weight_dead=0.0047775)
#assume dragonet deaths are approximately 1/4 of boarfish catches
yield<-rbind(yield,Red_gurnard,Boarfish, Dragonet)


total_weight_dead<-FDI_EU%>%
    left_join(yield, by='species_code')%>%
    mutate(total_weight_dead_gear_per_area=(proportion_of_total_weight_dead*total_weight_dead))
```

compare total weight dead species and gear categories, if no size distribution is present for a category add the deaths to misc.

```{r}
#Blue whiting
blue_whiting_distribution<-subset(Celtic_Sea_Fishing, species=="Blue whiting")
blue_whiting_deaths<-subset(total_weight_dead, species=="Blue whiting")
unique(blue_whiting_distribution$gear_category)
unique(blue_whiting_deaths$gear_category)
    #Gillnet not included in distribution, so place gillnet deaths into Mix
blue_whiting_deaths<-blue_whiting_deaths%>%
    mutate(gear_category = case_when(
        gear_category %in% c("Gillnet") ~ "Mix",
        TRUE ~ gear_category))%>%
    group_by(gear_category,species)%>%
    summarise(total_weight_dead_gear_per_area=sum(total_weight_dead_gear_per_area),
              .groups = "drop")

#Boarfish
boarfish_distribution<-subset(Celtic_Sea_Fishing, species=="Boarfish")
boarfish_deaths<-subset(total_weight_dead, species=="Boarfish")
unique(boarfish_distribution$gear_category)
unique(boarfish_deaths$gear_category)
    #Pelagic,Gillnet, and Misc not included in distribution, so place gillnet and pelagic deaths mix gear
boarfish_deaths<-boarfish_deaths%>%
    mutate(gear_category = case_when(
        gear_category %in% c("Pelagic") ~ "Mix",
        gear_category %in% c("Gillnet") ~ "Mix",
        TRUE ~ gear_category))%>%
    group_by(gear_category,species)%>%
    summarise(total_weight_dead_gear_per_area=sum(total_weight_dead_gear_per_area),
              .groups = "drop")
#Cod
cod_distribution<-subset(Celtic_Sea_Fishing, species=="Cod")
cod_deaths<-subset(total_weight_dead, species=="Cod")
unique(cod_distribution$gear_category)
unique(cod_deaths$gear_category)
    #Misc not included in distribution, so Misc deaths into mix gear
cod_deaths<-cod_deaths%>%
    mutate(gear_category = case_when(
        gear_category %in% c("Misc") ~ "Mix",
        TRUE ~ gear_category))%>%
    group_by(gear_category,species)%>%
    summarise(total_weight_dead_gear_per_area=sum(total_weight_dead_gear_per_area),
              .groups = "drop")

#Dragonet
dragonet_distribution<-subset(Celtic_Sea_Fishing, species=="Dragonet")
dragonet_deaths<-subset(total_weight_dead, species=="Dragonet")
unique(dragonet_distribution$gear_category)
unique(dragonet_deaths$gear_category)
    #Gillnet not included in distribution, so place gillnet and trawl deaths into mix
dragonet_deaths<-dragonet_deaths%>%
    mutate(gear_category = case_when(
        gear_category %in% c("Gillnet") ~ "Mix",
        TRUE ~ gear_category))%>%
    group_by(gear_category,species)%>%
    summarise(total_weight_dead_gear_per_area=sum(total_weight_dead_gear_per_area),
              .groups = "drop")

#Haddock
haddock_distribution<-subset(Celtic_Sea_Fishing, species=="Haddock")
haddock_deaths<-subset(total_weight_dead, species=="Haddock")
unique(haddock_distribution$gear_category)
unique(haddock_deaths$gear_category)
    #Misc and Pelagic not included in distribution, so place Misc and Pelagic deaths into mix
haddock_deaths<-haddock_deaths%>%
    mutate(gear_category = case_when(
        gear_category %in% c("Misc") ~ "Mix",
        gear_category %in% c("Pelagic") ~ "Mix",
        TRUE ~ gear_category))%>%
    group_by(gear_category,species)%>%
    summarise(total_weight_dead_gear_per_area=sum(total_weight_dead_gear_per_area),
              .groups = "drop")

#Hake
hake_distribution<-subset(Celtic_Sea_Fishing, species=="Hake")
hake_deaths<-subset(total_weight_dead, species=="Hake")
unique(hake_distribution$gear_category)
unique(hake_deaths$gear_category)
    #Misc not included in distribution, so place Misc deaths into mix
hake_deaths<-hake_deaths%>%
    mutate(gear_category = case_when(
        gear_category %in% c("Misc") ~ "Mix",
        TRUE ~ gear_category))%>%
    group_by(gear_category,species)%>%
    summarise(total_weight_dead_gear_per_area=sum(total_weight_dead_gear_per_area),
              .groups = "drop")

#Herring
herring_distribution<-subset(Celtic_Sea_Fishing, species=="Herring")
herring_deaths<-subset(total_weight_dead, species=="Herring")
unique(herring_distribution$gear_category)
unique(herring_deaths$gear_category)
    #Misc not included in distribution, so place Misc deaths into mix
herring_deaths<-herring_deaths%>%
    mutate(gear_category = case_when(
        gear_category %in% c("Misc") ~ "Mix",
        TRUE ~ gear_category))%>%
    group_by(gear_category,species)%>%
    summarise(total_weight_dead_gear_per_area=sum(total_weight_dead_gear_per_area),
              .groups = "drop")

#Horse mackerel
horse_mackerel_distribution<-subset(Celtic_Sea_Fishing, species=="Horse mackerel")
horse_mackerel_deaths<-subset(total_weight_dead, species=="Horse mackerel")
unique(horse_mackerel_distribution$gear_category)
unique(horse_mackerel_deaths$gear_category)
    #Gillnet and Misc not included in distribution, so place Misc and Gillnet deaths into mix
horse_mackerel_deaths<-horse_mackerel_deaths%>%
    mutate(gear_category = case_when(
        gear_category %in% c("Misc") ~ "Mix",
         gear_category %in% c("Gillnet") ~ "Mix",
        TRUE ~ gear_category))%>%
    group_by(gear_category,species)%>%
    summarise(total_weight_dead_gear_per_area=sum(total_weight_dead_gear_per_area),
              .groups = "drop")

#Mackerel
mackerel_distribution<-subset(Celtic_Sea_Fishing, species=="Mackerel")
mackerel_deaths<-subset(total_weight_dead, species=="Mackerel")
unique(mackerel_distribution$gear_category)
unique(mackerel_deaths$gear_category)
    #All gears match remove miz gear from mackerel
mackerel_distribution<-mackerel_distribution%>%
    filter(gear_category !="Mix")
mackerel_deaths<-mackerel_deaths%>%
    group_by(gear_category,species)%>%
    summarise(total_weight_dead_gear_per_area=sum(total_weight_dead_gear_per_area),
              .groups = "drop")

#Megrim
megrim_distribution<-subset(Celtic_Sea_Fishing, species=="Megrim")
megrim_deaths<-subset(total_weight_dead, species=="Megrim")
unique(megrim_distribution$gear_category)
unique(megrim_deaths$gear_category)
    #Gillnet, Pelagic, and Misc not included in distribution, so place Misc, Pelagic, and Gillnet deaths into mix
megrim_deaths<-megrim_deaths%>%
    mutate(gear_category = case_when(
        gear_category %in% c("Misc") ~ "Mix",
        gear_category %in% c("Gillnet") ~ "Mix",
        gear_category %in% c("Pelagic") ~ "Mix",
        TRUE ~ gear_category))%>%
    group_by(gear_category,species)%>%
    summarise(total_weight_dead_gear_per_area=sum(total_weight_dead_gear_per_area),
              .groups = "drop")

#Monkfish
monkfish_distribution<-subset(Celtic_Sea_Fishing, species=="Monkfish")
monkfish_deaths<-subset(total_weight_dead, species=="Monkfish")
unique(monkfish_distribution$gear_category)
unique(monkfish_deaths$gear_category)
    #Gillnet and Pelagic not included in distribution, so place Pelagic and Gillnet deaths into mix
monkfish_deaths<-monkfish_deaths%>%
    mutate(gear_category = case_when(
        gear_category %in% c("Gillnet") ~ "Mix",
        gear_category %in% c("Pelagic") ~ "Mix",
        TRUE ~ gear_category))%>%
    group_by(gear_category,species)%>%
    summarise(total_weight_dead_gear_per_area=sum(total_weight_dead_gear_per_area),
              .groups = "drop")

#Plaice
plaice_distribution<-subset(Celtic_Sea_Fishing, species=="Plaice")
plaice_deaths<-subset(total_weight_dead, species=="Plaice")
unique(plaice_distribution$gear_category)
unique(plaice_deaths$gear_category)
    #All gears shared, remove mix from distribution
plaice_distribution<-plaice_distribution%>%
    filter(gear_category !="Mix")
plaice_deaths<-plaice_deaths%>%
    group_by(gear_category,species)%>%
    summarise(total_weight_dead_gear_per_area=sum(total_weight_dead_gear_per_area),
              .groups = "drop")

#Red gurnard
red_gurnard_distribution<-subset(Celtic_Sea_Fishing, species=="Red gurnard")
red_gurnard_deaths<-subset(total_weight_dead, species=="Red gurnard")
unique(red_gurnard_distribution$gear_category)
unique(red_gurnard_deaths$gear_category)
    #Misc not included in distribution, so place Misc deaths into mix
red_gurnard_deaths<-red_gurnard_deaths%>%
    mutate(gear_category = case_when(
        gear_category %in% c("Misc") ~ "Mix",
        TRUE ~ gear_category))%>%
    group_by(gear_category,species)%>%
    summarise(total_weight_dead_gear_per_area=sum(total_weight_dead_gear_per_area),
              .groups = "drop")

#Sole
sole_distribution<-subset(Celtic_Sea_Fishing, species=="Sole")
sole_deaths<-subset(total_weight_dead, species=="Sole")
unique(sole_distribution$gear_category)
unique(sole_deaths$gear_category)
    #Pelagic not included in distribution, so place Pelagic deaths into mix
sole_deaths<-sole_deaths%>%
    mutate(gear_category = case_when(
        gear_category %in% c("Pelagic") ~ "Mix",
        TRUE ~ gear_category))%>%
    group_by(gear_category,species)%>%
    summarise(total_weight_dead_gear_per_area=sum(total_weight_dead_gear_per_area),
              .groups = "drop")

#Whiting
whiting_distribution<-subset(Celtic_Sea_Fishing, species=="Whiting")
whiting_deaths<-subset(total_weight_dead, species=="Whiting")
unique(whiting_distribution$gear_category)
unique(whiting_deaths$gear_category)
    #Misc not included in distribution, so place Misc deaths into mix
whiting_deaths<-whiting_deaths%>%
    mutate(gear_category = case_when(
        gear_category %in% c("Misc") ~ "Mix",
        TRUE ~ gear_category))%>%
    group_by(gear_category,species)%>%
    summarise(total_weight_dead_gear_per_area=sum(total_weight_dead_gear_per_area),
              .groups = "drop")


#combine for final distribution and deaths
catch_distribution<-rbind(blue_whiting_distribution,boarfish_distribution,cod_distribution,dragonet_distribution,haddock_distribution,hake_distribution,herring_distribution,horse_mackerel_distribution,mackerel_distribution,megrim_distribution,monkfish_distribution,plaice_distribution,red_gurnard_distribution,sole_distribution,whiting_distribution)

fishing_deaths<-rbind(blue_whiting_deaths,boarfish_deaths,cod_deaths,dragonet_deaths, haddock_deaths,hake_deaths,herring_deaths,horse_mackerel_deaths,mackerel_deaths, megrim_deaths,monkfish_deaths,plaice_deaths,red_gurnard_deaths,sole_deaths,whiting_deaths)

##
#use_data(catch_distribution, overwrite = TRUE)
#use_data(fishing_deaths, overwrite = TRUE)
```
